## Exercise (A) {.unnumbered}

::: callout-tip
You can download all of the related files in a zip file **accessingEx.zip** from [Github](https://github.com/ehsanx/EpiMethods/tree/main/LabExercises/)

-   Navigate to the GitHub page (above link) where the ZIP file is located.
-   Click on the file name (above zip file) to open its preview window.
-   Click on the Download button to download the file. If you can't see the Download button, click on "Download Raw File" link that should appear on the page.
:::

## Problem Statement

We will use the article by @palis2020relationship, [DOI: 10.1080/09638237.2018.1437602](https://www.researchgate.net/profile/Heather_Palis/publication/323136807_The_relationship_between_sense_of_community_belonging_and_self-rated_mental_health_among_Canadians_with_mental_or_substance_use_disorders/links/5a96ff210f7e9ba42974c0dd/The-relationship-between-sense-of-community-belonging-and-self-rated-mental-health-among-Canadians-with-mental-or-substance-use-disorders.pdf).

-   Download the [CCHS MH topical index](https://gsg.uottawa.ca/data/teaching/escc-cchs/fmgd_pumf/2012/cchs_mh_topical_index.pdf)
-   Download the [CCHS MH Data Dictionary](https://gsg.uottawa.ca/data/teaching/escc-cchs/fmgd_pumf/2012/cchs_mh_datadictionary.pdf)

## Question I: \[60% grade\]

### 1(a) Importing dataset

```{r}
# Importing dataset
load("Data/accessing/cchsMH.RData") 
```

### 1(b) Subsetting according to eligibility

Subset the dataset according to the eligibility criteria / restriction specified in the paper

-   Identify the variable needed for eligibility criteria

-   Identify the outcome variable

-   Identify the explanatory variable

-   Identify the potential confounders

-   Identify the survey weight variable

-   **Hint**

    1.  Read

    -   the first paragraph of **Analytic sample** (page 2) for the eligibility criteria, and
    -   first and second paragraphs of **Study variables** for rest of the variables,
    -   third paragraph of the **Statistical analyses** for the survey weights variable.

    2.  eligibility criteria was determined based on **one** variable. Only work with 'YES' category.
    3.  Outcome variable has a category 'NOT STATED', but for our analysis, we will omit anyone associated with this category.
    4.  For explanatory variable, we have categories such as DON'T KNOW, REFUSAL and NOT STATED. We will not use these categories (omit anyone with these categories).
    5.  There were **five** potential confounders.
    6.  Potentially useful functions:

    -   [%in%](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/match)
    -   [levels](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/levels)
    -   [recode](https://rdrr.io/cran/car/man/recode.html)
    -   [subset](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/subset)
    -   [as.factor](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/factor)
    -   [relevel](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/relevel)

    or [dplyr ways](https://ehsanx.github.io/intro2R/data-analysis-with-dplyr.html):

    -   [filter](https://www.rdocumentation.org/packages/dplyr/versions/0.7.8/topics/filter)
    -   [select](https://www.rdocumentation.org/packages/dplyr/versions/0.7.8/topics/select)

```{r}
# your code here

```

### 1(c) Retaining necessary variables

In the dataset, retain only the variables associated with outcome measure, explanatory variable, potential confounders and survey weight

```{r}
# your code here

```

### 1(d) Creating analytic dataset

-   Assign missing values for categories such as DON'T KNOW, REFUSAL and NOT STATED.
-   'recode' the variables as shown in Table 1 (choose any function of your choice). Here is an example (but feel free to use other functions. In R there are many other ways to do this same task):

```{r}
## your code here
# levels(your.data.frame$your.age.variable) <- 
#   list("15 to 24 years" = c("15 TO 19 YEARS", "20 TO 24 YEARS"),
#        "25 to 34 years" = c("25 TO 29 YEARS", "30 TO 34 YEARS"),
#        "35 to 44 years" = c("35 TO 39 YEARS", "40 TO 44 YEARS"),
#        "45 to 54 years" = c("45 TO 49 YEARS", "50 TO 54 YEARS"),
#        "55 to 64 years" = c("55 TO 59 YEARS", "60 TO 64 YEARS"),
#        "65 years or older" = c("65 TO 69 YEARS", "70 TO 74 YEARS", 
#        "75 TO 79 YEARS", "80 YEARS OR MORE"))
```

### 1(e) Number of columns and variable names

report the number of columns in your analytic dataset, and the variable names.

```{r}
# your code here

```

## Question II: \[20% grade\]

### 2(a) Table 1

Reproduce Table 1 presented in the above paper (omit the 'Main source of income' variable). If necessary, drop other variables from the analytic dataset that are not presented in Table 1.

The table you produce should report numbers as follows:

| Self-rated Mental Health Variable | Total n(%)  | Poor or Fair n(%) | Good n(%)   | Very good or excellent n(%) |
|-----------------------------------|-------------|-------------------|-------------|-----------------------------|
| **Study sample**                  | 2628 (100)  | 1002 (38.1)       | 885 (33.7)  | 741 (28.2)                  |
| **Community belonging**           |             |                   |             |                             |
| \- Very weak                      | 480 (18.3)  | 282 (28.1)        | 118 (13.3)a | 80 (10.8)a                  |
| \- Somewhat weak                  | 857 (32.6)  | 358 (35.7)        | 309 (34.9)  | 190 (25.6)                  |
| \- Somewhat strong                | 1005 (38.2) | 288 (28.7)        | 362 (40.9)  | 355 (47.9)                  |
| \- Very strong                    | 286 (10.9)  | 74 (7.4)a         | 96 (10.8)a  | 116 (15.7)a                 |
| **Sex**                           |             |                   |             |                             |
| \- Females                        | 1407 (53.5) | 616 (61.5)        | 487 (55.0)  | 304 (41.0)                  |
| \- Males                          | 1221 (46.5) | 386 (38.5)        | 398 (45.0)  | 437 (59.0)                  |
| **Age group**                     |             |                   |             |                             |
| \- 15 to 24 years                 | 740 (28.2)  | 191 (19.1)        | 264 (29.8)  | 285 (38.5)                  |
| \- 25 to 34 years                 | 475 (18.1)  | 141 (14.1)        | 167 (18.9)  | 167 (22.5)                  |
| \- 35 to 44 years                 | 393 (15.0)  | 185 (18.5)        | 119 (13.4)a | 89 (12.0)a                  |
| \- 45 to 54 years                 | 438 (16.6)  | 220 (22.0)        | 139 (15.7)  | 79 (10.7)a                  |
| \- 55 to 64 years                 | 379 (14.4)  | 198 (19.7)        | 113 (12.8)a | 68 (9.2)a                   |
| \- 65 years or older              | 203 (7.7)   | 67 (6.6)a         | 83 (8.4)a   | 53 (7.1)b                   |
| **Race/Ethnicity**                |             |                   |             |                             |
| \- Non-white                      | 458 (17.4)  | 184 (18.4)        | 140 (15.8)  | 134 (18.1)                  |
| \- White                          | 2170 (82.6) | 818 (81.6)        | 745 (84.2)  | 607 (81.9)                  |
| **Main source of income**         |             |                   |             |                             |
| \- Employment Income\^d           | 1054 (40.1) | 289 (28.8)        | 386 (43.6)  | 379 (51.1)                  |
| \- Worker's Compensation\^e       | 160 (6.1)   | 91 (9.1)a         | 44 (5.0)b   | 25 (3.4)c                   |
| \- Senior Benefits\^f             | 134 (5.1)   | 57 (5.7)a         | 42 (4.7)b   | 35 (4.7)                    |
| \- Other\^g                       | 184 (7.0)   | 82 (8.2)a         | 60 (6.8)a   | 42 (5.7)b                   |
| \- Not applicable\^h              | 851 (32.4)  | 402 (40.1)        | 263 (29.7)  | 186 (25.1)                  |
| \- Not Stated\^i                  | 245 (9.3)   | 81 (8.1)a         | 90 (10.2)a  | 74 (10.0)                   |

$^a$ Coefficient of variation between 16.6 and 25.0%. $^b$ Coefficient of variation between 25.1 and 33.3%. $^c$ Coefficient of variation \> 33.3%. $^d$ Employment Income: Wages/salaries or self-employment. $^e$ Worker's compensation: Employment insurance or worker's compensation or social assistance/welfare. $^f$ Senior Benefits: Benefits from Canada or Quebec Pension Plan or job related retirement pensions, superannuation and annuities or RRSP/RRIF of Old Age Security and Guaranteed Income Supplement. $^g$ Other: Dividends/interest or child tax benefit or child support or alimony or other or no income. $^h$ Not applicable: Respondents who live in a household with only one person. The income variable "main source of personal income" is applicable only to those that live in a household of more than one person. $^i$ Not Stated: Question was not answered (don't know, refusal, not stated). - **Hint** - You can produce 1 table with total, and another table stratified by the necessary variable.

```{r,warning=FALSE,message=FALSE}
# your code here
require(tableone)

```

## Question III: \[20% grade\]

### 3(a) Subset

Subset the dataset excluding 'Very good or excellent' responses from the self-rated mental health variable

```{r}
# your code here

```

### 3(b) Recode

Recode self-rated mental health variable and make it a binary variable: 'Good' vs. 'Poor' (simplifying category labels only). Convert that variable to a factor variable with 'Poor' being the reference level.

```{r}
# your code here

```

### 3(c) Regression

Run a logistic regression model for finding the relationship between community belonging (Reference: Very weak) and self-rated mental health (Reference: Poor) among respondents with mental or substance use disorders. Adjust the model for three confounders: sex, age, and race/ethnicity.

```{r}
# your code here

```

### 3(d) Reporting odds ratio

Report the odds ratios and associated confidence intervals (use Publish package).

```{r,warning=FALSE,message=FALSE}
require(Publish)
# your code here

```
