# The Project {.unnumbered}

Welcome to a place crafted to bridge a unique gap in the health research world. This website offers valuable resources for those who are taking their first steps into health research and advanced statistics. Even if you're familiar with health research, but advanced statistical methods seem daunting, you're in the right place. Here, we offer:

-   Tutorials on using the R software, starting from the basics.
-   Guides on tapping into genuine survey data from reputable sources like Statistics Canada and the US government's CDC.
-   Step-by-step walkthroughs on conducting and reporting comprehensive epidemiological studies.

This hub is a part of an open educational initiative, meaning it's available to everyone. We hope to uplift the standard of health research methodology through this endeavor.

# What We Aim to Achieve {.unnumbered}

We're on a mission to:

-   Equip public health learners with hands-on experience.
-   Teach the nuances of applying advanced epidemiological methods using real data.
-   Offer a unique open textbook that's enriched with interactive tools and quizzes for a self-paced learning experience.

## Dive into Our Modules {.unnumbered}

Embark on a journey through 10 core learning modules, including one introductory module about R:

```{r, echo=FALSE, warning=FALSE,message=FALSE}
library(knitr)
library(kableExtra)
topics <- c("R Data Wrangling", "Crafting Data for Research", "Survey Data Resources", "Grasping Confounding", "Adjustment Techniques", "Complex Survey Analysis", "Missing Data", "Propensity Score Analysis", "Reporting Guideline", "Machine Learning", "Machine Learners in Causal Inference")
descriptions <- c("Get to know R.", "Customize data to your research query.", "Understand and source reliable national survey data.", "Delve into the concept of confounding and its implications.", "Adjusting for covariates in a multivariate analysis.", "Handle intricate data sets.", "Understand and tackle gaps in your data.", "Dive deeper into advanced analyses.", "Guidelines to share your findings.", "Introduction to key concepts, algorithms, and applications.", "Discusses the potential pitfalls and challenges in merging machine learning with causal inference, and a way forward.")
df <- data.frame(`Module` = 0:10, Topics = topics, Descriptions = descriptions)
table_content <- kable(df, "html", align = c("c", "l", "l"), escape = FALSE)
table_content <- gsub("R Data Wrangling", "<a href='wrangling.html'>R Data Wrangling</a>", table_content)
table_content <- gsub("Survey Data Resources", "<a href='accessing.html'>Survey Data Resources</a>", table_content)
table_content <- gsub("Crafting Data for Research", "<a href='researchquestion.html'>Crafting Data for Research</a>", table_content)
table_content <- gsub("Grasping Confounding", "<a href='confounding.html'>Grasping confounding</a>", table_content)
table_content <- gsub("Adjustment Techniques", "<a href='predictivefactors.html'>Adjustment Techniques</a>", table_content)
table_content <- gsub("Complex Survey Analysis", "<a href='surveydata.html'>Complex Survey Analysis</a>", table_content)
table_content <- gsub("Missing Data", "<a href='missingdata.html'>Missing Data</a>", table_content)
table_content <- gsub("Propensity Score Analysis", "<a href='propensityscore.html'>Propensity Score Analysis</a>", table_content)
table_content <- gsub("Reporting Guideline", "<a href='reporting.html'>Reporting Guideline</a>", table_content)
table_content <- gsub("Machine Learning", "<a href='machinelearning.html'>Machine Learning</a>", table_content)
table_content <- gsub("Machine Learners in Causal Inference", "<a href='machinelearningCausal.html'>Machine Learners in Causal Inference</a>", table_content)
table_content %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

## How Our Content is Presented {.unnumbered}

All our resources are hosted on an easy-to-access GitHub page. The format? Engaging text, reproducible software codes, clear analysis outputs, and crisp videos that distill complex topics. And don't miss our quiz section at the end of each module for a quick self-check on what you've learned.

## Open Copyright License {.unnumbered}

CC-BY

## Contributor list

::: callout-tip
Dive into this captivating content, brought to life with the generous support of the [UBC OER Fund Implementation Grant](https://oerfund.open.ubc.ca/oer-implementation-grants) and further supported by [UBC SPPH](https://spph.ubc.ca/). The foundation of this content traces back to the PI's work over five years while instructing [SPPH 604](https://spph.ubc.ca/graduate-courses/) (2018-2022). That knowledge have now transformed into an open educational resource, thanks to this grant. Meet the innovative minds behind the grant proposal below.
:::

```{r, echo=FALSE, warning=FALSE,message=FALSE}
library(knitr)
library(kableExtra)

data <- data.frame(
  Role = c("Principal Applicant (PI)", "Co-applicant (Co-I)", "Trainee co-applicants", "", "", "", ""),
  Team_Member = c("Mohammad Ehsanul Karim", "Dr. Suborna Ahmed", "Belal Hossain", "Fardowsa Yusuf", "Hanna Frank", "Dr. Michael Asamoah-Boaheng", "Chuyi (Astra) Zheng"),
  Affiliation = c("UBC Department of Forest Resources Management", "UBC Department of Forest Resources Management", "UBC School of Population and Public Health", "UBC School of Population and Public Health", "UBC School of Population and Public Health", "UBC Department of Emergency Medicine", "UBC Faculty of Arts")
)

kable(data, "html", escape = F) %>%
  kable_styling("striped", full_width = F) %>%
  row_spec(0, background = "black", color = "white", bold = T) %>%
  row_spec(1, background = "white") %>%
  row_spec(2, background = "white") %>%
  row_spec(3:7, background = "white")

```
