# The Project {.unnumbered}

Welcome to a place crafted to bridge a unique gap in the health research world. This website offers valuable resources for those who are taking their first steps into health research and advanced statistics. Even if you're familiar with health research, but advanced statistical methods seem daunting, you're in the right place. Here, we offer:

-   Tutorials on using the R software, starting from the basics.
-   Guides on tapping into genuine survey data from reputable sources like Statistics Canada and the US government's CDC.
-   Step-by-step walkthroughs on conducting and reporting comprehensive epidemiological studies.

This hub is a part of an open educational initiative, meaning it's available to everyone. We hope to uplift the standard of health research methodology through this endeavor.

## What We Aim to Achieve {.unnumbered}

We're on a mission to:

-   Equip public health learners with hands-on experience.
-   Teach the nuances of applying advanced epidemiological methods using real data.
-   Offer a unique open textbook that's enriched with interactive tools and quizzes for a self-paced learning experience.

## Dive into Our Modules {.unnumbered}

Embark on a journey through 10 core learning modules, and one introductory module about R (module 0). Letters in the parentheses (W, A, Q, R, P, D, M, S, L, C, G) are the chapter indicators; these are indicated along with quizzes, R functions, and exercises associated with the corresponding chapters. Only key chapters have exercises.

```{r, echo=FALSE, warning=FALSE,message=FALSE}
library(knitr)
library(kableExtra)
topics <- c("R Data Wrangling", "Survey Data Resources", "Crafting Data for Research", "Grasping Confounding", "Adjustment Techniques", "Complex Survey Analysis", "Missing Data", "Propensity Score Analysis", "Machine Learning", "Machine Learners in Causal Inference", "Non-binary outcomes", "Longitudinal analysis", "Mediation analysis", "Writing Tools")
descriptions <- c("Get to know R.", "Understand and source reliable national survey data.", "Customize data to your research query.", "Delve into the concept of confounding and its implications.", "Introduction to key concepts of prediction modelling.", "Handle data sets obtained from complex survey designs.", "Understand and tackle missingness in your data.", "Dive deeper into advanced observational data analyses.", "Introduction to machine learning algorithms, and applications.", "Discusses the potential pitfalls and challenges in merging machine learning with causal inference, and a way forward.", "Statistical techniques to deal with complex or non-binary outcomes", "Longitudinal data analysis techniques", "Mediation: decomposing the total effect", "Tools and guides for scientific writing and collaboration.")
df <- data.frame(`Module` = 0:13, 'Topics-Indicators' = topics, Descriptions = descriptions)
table_content <- kable(df, "html", align = c("c", "l", "l"), escape = FALSE)
table_content <- gsub("R Data Wrangling", "<a href='wrangling.html'>R for Data Wrangling (W)</a>", table_content)
table_content <- gsub("Survey Data Resources", "<a href='accessing.html'>Accessing (A) Survey Data Resources</a>", table_content)
table_content <- gsub("Crafting Data for Research", "<a href='researchquestion.html'>Crafting Analytic Data for Research Questions (Q)</a>", table_content)
table_content <- gsub("Grasping Confounding", "<a href='confounding.html'>Causal roles (R)</a>", table_content)
table_content <- gsub("Adjustment Techniques", "<a href='predictivefactors.html'>Predictive (P) issues</a>", table_content)
table_content <- gsub("Complex Survey Analysis", "<a href='surveydata.html'>Complex Survey Data (D) Analysis</a>", table_content)
table_content <- gsub("Missing Data", "<a href='missingdata.html'>Missing (M) Data Analysis</a>", table_content)
table_content <- gsub("Propensity Score Analysis", "<a href='propensityscore.html'>Propensity Score (S) Analysis</a>", table_content)
table_content <- gsub("Writing Tools", "<a href='reporting.html'>Scientific Writing Tools (G)</a>", table_content)
table_content <- gsub("Machine Learning", "<a href='machinelearning.html'>Machine Learning (L)</a>", table_content)
table_content <- gsub("Machine Learners in Causal Inference", "<a href='machinelearningCausal.html'>Intergrating Machine Learners in Causal (C) Inference</a>", table_content)
table_content <- gsub("Non-binary outcomes", "<a href='nonbinary.html'>Non-binary outcomes (N)</a>", table_content)
table_content <- gsub("Longitudinal analysis", "<a href='longitudinal.html'>Longitudinal analysis (T)</a>", table_content)
table_content <- gsub("Mediation analysis", "<a href='mediation.html'>Mediation analysis (I)</a>", table_content)
table_content %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

::: callout-note
The tutorial is designed with a consistent structure across all chapters to provide a cohesive and thorough learning experience. Here's what you can expect in each chapter:

1.  **Overview**: The first page of each chapter offers a concise summary that outlines the key learning objectives, topics covered, and what you can expect to gain from the chapter. The overview page will also feature links to the data sources used in the tutorials as well as a [form](https://forms.gle/YSwuiebtb1E9wjHu9) where you can report any bugs or issues you encounter. This helps you quickly grasp the chapter's essence and set learning expectations.

2.  **Tutorial Topics**: Immediately following the overview, you'll find in-depth tutorials that cover each topic in detail. These are designed to provide comprehensive insights and are spread across multiple pages for easier navigation and understanding.

3.  **Summary of R Functions**: Each chapter includes a succinct summary of the R functions used in the tutorials. This serves as a quick reference guide for learners to understand the tools they will be applying.

4.  **Chapter-Specific Quiz**: For those interested in self-assessment, each chapter concludes with an optional quiz. This is a self-paced learning tool to help reinforce the chapter's key concepts.

5.  **Practice Exercises**: Finally, practice exercises are available for selected chapters to help you apply what you've learned in a hands-on manner. These exercises are designed to reinforce your understanding and give you practical experience with the chapter's topics.
:::

## How Our Content is Presented {.unnumbered}

All our resources are hosted on an easy-to-access GitHub page. The format? Engaging text, reproducible software codes, clear analysis outputs, and crisp videos that distill complex topics. And don't miss our quiz section at the end of each module for a quick self-check on what you've learned. This document is created using [quarto](https://quarto.org/) and [R](https://cran.r-project.org/).

## Open Copyright License {.unnumbered}

CC-BY

## Contributor list

Dive into this captivating content, brought to life with the generous support of the [UBC OER Fund Implementation Grant](https://oerfund.open.ubc.ca/oer-implementation-grants) and further supported by [UBC SPPH](https://spph.ubc.ca/). The foundation of this content traces back to the [PI](https://ehsank.com/)'s work over five years while instructing [SPPH 604](https://spph.ubc.ca/graduate-courses/) (2018-2022). That knowledge have now transformed into an open educational resource, thanks to this grant. Meet the innovative minds behind the grant proposal below.

```{r, echo=FALSE, warning=FALSE,message=FALSE}
library(knitr)
library(kableExtra)

data <- data.frame(
  Role = c("Principal Applicant (PI)", "Co-applicant (Co-I)", "Trainee co-applicants", "", "", "", ""),
  Team_Member = c("Dr. M Ehsan Karim", "Dr. Suborna Ahmed", "Md Belal Hossain", "Fardowsa Yusuf", "Hanna Frank", "Dr. Michael Asamoah-Boaheng", "Chuyi (Astra) Zheng"),
  Affiliation = c("UBC School of Population and Public Health", "UBC Department of Forest Resources Management", "UBC School of Population and Public Health", "UBC School of Population and Public Health", "UBC School of Population and Public Health", "UBC Department of Emergency Medicine", "UBC Faculty of Arts")
)

kable(data, "html", escape = F) %>%
  kable_styling("striped", full_width = F) %>%
  row_spec(0, background = "black", color = "white", bold = T) %>%
  row_spec(1, background = "white") %>%
  row_spec(2, background = "white") %>%
  row_spec(3:7, background = "white")

```

Additional earlier contributors to the course material development, who were not part of this current OER grant, include Derek Ouyang, Kate McLeod (both from UBC School of Population and Public Health), and Mohammad Atiquzzaman (UBC Pharmaceutical Sciences). Numerous pieces of student feedback were also incorporated in order to update the content.
