## Binary outcome {.unnumbered}

::: callout-important
This tutorial is very similar to one of the [previous tutorials](predictivefactors3.html), but uses a different data (we used [RHC data](researchquestion1.html) here). We are revisiting concepts related to [prediction](predictivefactors.html) before introducing ideas related to machine learning. 
:::

In this chapter, we will talk about Regression that deals with prediction of binary outcomes. We will use logistic regression to build the first prediction model.

```{r setup05, include=FALSE}
require(Publish)
```

### Read previously saved data

```{r load, cache=TRUE}
ObsData <- readRDS(file = "Data/machinelearning/rhcAnalytic.RDS")
```

### Outcome levels (factor)

-   Label
    -   Possible values of outcome

```{r levels, cache=TRUE}
levels(ObsData$Death)=c("No","Yes") # this is useful for caret
# ref: https://tinyurl.com/caretbin
class(ObsData$Death)
table(ObsData$Death)
```

### Measuring prediction error

-   Brier score
    -   Brier score 0 means perfect prediction, and
    -   close to zero means better prediction,
    -   1 being the worst prediction.
    -   Less accurate forecasts get higher score in Brier score.
-   AUC
    -   The area under a ROC curve is called as a c statistics.
    -   c being 0.5 means random prediction and
    -   1 indicates perfect prediction

### Prediction for death

In this section, we show the regression fitting when outcome is binary (death).

#### Variables

```{r vars, cache=TRUE, echo = TRUE}
baselinevars <- names(dplyr::select(ObsData, 
                         !c(Length.of.Stay,Death)))
baselinevars
```

#### Model

```{r reg3, cache=TRUE, echo = TRUE, results='hide', warning=FALSE}
# adjust covariates
out.formula2 <- as.formula(paste("Death~ ", paste(baselinevars, collapse = "+")))
saveRDS(out.formula2, file = "Data/machinelearning/form2.RDS")
fit2 <- glm(out.formula2, data = ObsData, 
            family = binomial(link = "logit"))
require(Publish)
adj.fit2 <- publish(fit2, digits=1)$regressionTable
```

```{r reg3a, cache=TRUE, echo = TRUE}
out.formula2
adj.fit2
```

#### Measuring prediction error

##### AUC

```{r auc1, cache=TRUE, warning=FALSE}
require(pROC)
obs.y2<-ObsData$Death
pred.y2 <- predict(fit2, type = "response")
rocobj <- roc(obs.y2, pred.y2)
rocobj
plot(rocobj)
auc(rocobj)
```

##### Brier Score

```{r brier, cache=TRUE, warning=FALSE}
require(DescTools)
BrierScore(fit2)
```

### Cross-validation using caret

### Basic setup

```{r cvbin, cache= TRUE, warning=FALSE}
# Using Caret package
set.seed(504)

# make a 5-fold CV
require(caret)
ctrl<-trainControl(method = "cv", number = 5, 
                   classProbs = TRUE,
                   summaryFunction = twoClassSummary)

# fit the model with formula = out.formula2
# use training method glm (have to specify family)
fit.cv.bin<-train(out.formula2, trControl = ctrl,
               data = ObsData, method = "glm",
              family = binomial(),
              metric="ROC")
fit.cv.bin
```

#### Extract results from each test data

```{r cvbin2, cache= TRUE}
summary.res <- fit.cv.bin$resample
summary.res
mean(fit.cv.bin$resample$ROC)
sd(fit.cv.bin$resample$ROC)
```

#### More options

```{r cvbincs, cache= TRUE}
ctrl<-trainControl(method = "cv", number = 5, 
                   classProbs = TRUE,
                   summaryFunction = twoClassSummary)
fit.cv.bin<-train(out.formula2, trControl = ctrl,
               data = ObsData, method = "glm",
              family = binomial(),
              metric="ROC",
              preProc = c("center", "scale"))
fit.cv.bin
```

### Variable selection

We can also use stepwise regression that uses AIC as a criterion.

```{r cvbinaic, cache= TRUE, results = 'hide', warning=FALSE, message=FALSE, error=FALSE}
set.seed(504)
ctrl<-trainControl(method = "cv", number = 5, 
                   classProbs = TRUE,
                   summaryFunction = twoClassSummary)
fit.cv.bin.aic<-train(out.formula2, trControl = ctrl,
               data = ObsData, method = "glmStepAIC",
               direction ="backward",
              family = binomial(),
              metric="ROC")
```

```{r cvbinaic2, cache= TRUE}
fit.cv.bin.aic
summary(fit.cv.bin.aic)
```


### Video content (optional)

::: callout-tip
For those who prefer a video walkthrough, feel free to watch the video below, which offers a description of an earlier version of the above content.
:::

::: {style="position: relative; padding-bottom: 56.25%; height: 0; overflow: hidden;"}
<iframe src="https://www.youtube.com/embed/YPOP82Wxfy8" style="position: absolute; top: 0; left: 0; width: 100%; height: 100%; border:0;" allowfullscreen>

</iframe>
:::