## Survival analysis: NHANES {.unnumbered}

The tutorial demonstrates analyzing complex survey data (NHANES) with mortality as a survival outcome. See the [tutorial in the previous chapter](accessing7.html) on linking public-use US mortality data with the NHANES.

We will explore the relationship between caffeine consumption and mortality in adults with diabetes using NHANES 1999-2010 datasets. We will follow the following article by [Neves et al. (2018)](https://doi.org/10.3389/fendo.2018.00547).

### Load required packages

```{r setup, warning=FALSE, message=FALSE, cache=TRUE}
# Load required packages
library(tableone)
library(survival)
library(Publish)
library(survey)
library(DataExplorer)
```

### Load datasets

Let us load NHANES 1999-2000, 2001-2002, 2003-2004, 2005-2006, 2007-2008, and 2009-2010 datasets and merged all cycles.

```{r load, cache=TRUE, warning=FALSE, message=FALSE}
# Load
load("Data/nonbinary/coffee.RData")
ls()

# Merge all cycles
dat.full <- rbind(dat.analytic1999, dat.analytic2001, dat.analytic2003, 
                  dat.analytic2005, dat.analytic2007, dat.analytic2009)
head(dat.full)
dim(dat.full)
```

The merged dataset contains 62,160 subjects with 26 relevant variables:

-   id: Respondent sequence/ID number
-   survey.weight: Full sample 2 year weights
-   psu: Masked pseudo-PSU
-   strata: Masked pseudo-stratum
-   caff: Caffeine (exposure variable)
-   stime: Follow-up time (time from interview date to death or censoring)
-   status: Mortality status
-   sex: Sex
-   age: Age in years
-   race: Race/ethnicity
-   income: Annual household income
-   smoking: Smoking status
-   diabetic.nephropathy: Diabetic nephropathy (no data)
-   bmi.cat: BMI - categorical
-   education: Education level
-   carbohyd: Carbohydrate in gm
-   alcohol: Alcohol consumption
-   diab.years: Years since diabetes
-   htn: Hypertension
-   daib.retino: Diabetes retinopathy
-   macrovascular: Macrovascular complications
-   insulin: Insulin
-   survey.cycle: Survey cycle
-   physical.activity: Physical activity
-   diabetes: Diabetes status
-   cal.total: Total calories in kcal

### Data pre-processing

#### Eligibility criteria

The authors considered adults aged 18 years or more, only diabetic, and total calories between 500 and 3500.

```{r Eligibility, cache=TRUE, warning=FALSE, message=FALSE}
# Total samples in the merged dataset
nrow(dat.full) # N = 62,160

# Age >= 18 years
dat2 <- subset(dat.full, age >= 18) 
nrow(dat2) # N = 35,379

# With diabetes
dat3 <- subset(dat2, diabetes == "Yes") 
nrow(dat3) # N = 4,687 - numbers don't match with the paper (N = 4,544)

# Implausible alimentary reports
dat4 <- subset(dat3, cal.total >= 500 & cal.total <= 3500) 
nrow(dat4) # N = 4,083 - numbers don't match with the paper (N = 3,948)
```

#### Complete case data

Let us drop missing values in the exposure and outcome:

```{r cc1, cache=TRUE, warning=FALSE, message=FALSE}
# Drop missing exposure and outcome
dat <- dat4[complete.cases(dat4$id),]
dat <- dat[complete.cases(dat$caff),]
dat <- dat[complete.cases(dat$status),]
dat <- dat[complete.cases(dat$stime),] # N = 4,080
dim(dat)

# Missing plot
plot_missing(dat)
```

Now, let us drop variables with high missingness for this exercise. As explained in the [Missing data analysis chapter](missingdata.html), a better approach could be imputing missing values under the missing at random assumption.

```{r cc2, cache=TRUE, warning=FALSE, message=FALSE}
# Drop variables with high missingness
dat$diabetic.nephropathy <- dat$diab.years <- dat$daib.retino <- dat$income <- NULL

# Complete case data
dat <- na.omit(dat)
dim(dat) # N = 3,780
```

### Table 1

Now, let us create Table 1 stratified by coffee consumption (exposure), separately for males and females, as done in the article.

```{r tab1, cache=TRUE, warning=FALSE, message=FALSE}
vars <- c("age", "race", "education", "smoking", "alcohol", "carbohyd", "physical.activity",
          "bmi.cat", "htn", "macrovascular", "insulin", "survey.cycle")

tab1a <- CreateTableOne(vars = vars, strata = "caff", data = dat[dat$sex=="Female",], 
                        test = F)
tab1b <- CreateTableOne(vars = vars, strata = "caff", data = dat[dat$sex=="Male",], 
                        test = F)

tab1 <- list(Female = tab1a, Male = tab1b)
print(tab1, showAllLevels = T, smd = T)
```

### Survey design

The paper analyzed the data separately for males and females. Let us create the survey design:

```{r , cache=TRUE, warning=FALSE, message=FALSE}
# Revised weight - weight divided by 6 cycles
dat.full$svy.weight <- dat.full$survey.weight/6 
dat$svy.weight <- dat$survey.weight/6 
summary(dat$svy.weight)

# Create an indicator variable
dat.full$miss <- 1
dat.full$miss[dat.full$id %in% dat$id] <- 0

# Set up the design
w.design0 <- svydesign(strata = ~strata, id = ~psu, weights = ~svy.weight, 
                       data = dat.full, nest = TRUE)

# Subset the design
w.design1 <- subset(w.design0, miss == 0)

# Subset the design for females
w.design.f <- subset(w.design1, sex == "Female")
dim(w.design.f)

# Subset the design for males
w.design.m <- subset(w.design1, sex == "Male")
dim(w.design.m)
```

### Kaplan-Meier plot

Let us create the Kaplan-Meier plot for males and females:

```{r kmplot, cache=TRUE, warning=FALSE, message=FALSE}
# KM for females
fit0.f <- svykm(Surv(stime, status) ~ caff, design = w.design.f)
plot(fit0.f)

# KM for males
fit0.m <- svykm(Surv(stime, status) ~ caff, design = w.design.m)
plot(fit0.m)
```

### Cox PH

Now, we will fit the Cox proportional hazards model, separately for males and females. Let us run the unadjusted model first.

```{r coxunadj, cache=TRUE, warning=FALSE, message=FALSE}
# Unadjusted Cox PH for females
fit1.f <- svycoxph(Surv(stime, status) ~ caff, design = w.design.f)
publish(fit1.f)

# Unadjusted Cox PH for males
fit1.m <- svycoxph(Surv(stime, status) ~ caff, design = w.design.m)
publish(fit1.m)
```

Now, we will fit the Cox PH model, adjusting for covariates.

```{r coxadj, cache=TRUE, warning=FALSE, message=FALSE}
# Covariate adjusted Cox PH for females
fit2.f <- svycoxph(Surv(stime, status) ~ caff + age + race + education + smoking + 
                     alcohol +  carbohyd + physical.activity + bmi.cat + htn +
                     macrovascular + insulin + survey.cycle, design = w.design.f)
publish(fit2.f)

# Covariate adjusted Cox PH for males
fit2.m <- svycoxph(Surv(stime, status) ~ caff + age + race + education + smoking + 
                     alcohol +  carbohyd + physical.activity + bmi.cat + htn +
                     macrovascular + insulin + survey.cycle, design = w.design.m)
publish(fit2.m)
```

The adjusted results are approximately the same as in Table 2 of the article. Caffeine consumption was associated with mortality among women but not among men.

### PH assumption

Now, we will check the proportional hazard assumption.

```{r proportional, cache=TRUE, warning=FALSE, message=FALSE}
# PH assumption among females
cox.zph(fit2.f)

# PH assumption among males
cox.zph(fit2.m)
```

The large p-values indicate that the proportional hazard assumption was met for both models.
