## Poisson {.unnumbered}

In this tutorial, we will see how to use Poisson and negative binomial regression. In practice, we use Poisson regression to model a count outcome. Note that we assume the mean is equal to the variance in Poisson. When the variance is greater than what's assumed by the model, overdispersion occurs. Poisson regression of overdispersed data leads to under-estimated or deflated standard errors, which leads to inflated test statistics and overly small (anti-conservative) p-values, increasing false positives. We can use negative binomial regression to model overdispersed data.

```{r setup, warning=FALSE, message=FALSE, cache=TRUE}
# Load required packages
knitr::opts_chunk$set(echo = TRUE)
require(Publish)
require(survey)
```

### Data

```{r load, cache=TRUE}
# Load
load("Data/nonbinary/OAcvd.RData")

# Survey weight
analytic2$weight <- analytic2$weight/3

# Make fruit.cont as a numeric variable
analytic2$fruit.cont <- as.numeric(as.character(analytic2$fruit.cont))

# Make fruit.cont as a integer/count variable
analytic2$fruit.cont <- floor(analytic2$fruit.cont) # round

# Factor variables using lapply
var.names <- c("age", "sex", "income", "race", "bmicat", "phyact", "smoke",
               "fruit", "painmed", "ht", "copd", "diab", "edu", "CVD", "OA")
analytic2[var.names] <- lapply(analytic2[var.names] , factor)
str(analytic2)
```

### Survey weighted summary

```{r summary, cache=TRUE}
# Survey design
w.design <- svydesign(id=~1, weights=~weight, data=analytic2)

# Cross-tabulation
xtabs(~fruit.cont, analytic2)
svytable(~fruit.cont, w.design)
svyhist(~fruit.cont, w.design)
svyby(~fruit.cont, ~phyact, w.design, svymean, deff = TRUE)
```

### Poisson regression

Let us fit the traditional (not design-adjusted) Poisson regression using the `glm` function:

```{r pr, cache=TRUE}
require(jtools)
analytic2$phyact2=relevel(analytic2$phyact, ref ="Inactive")

# Poisson regression - crude
fit1 <- glm(fruit.cont ~phyact2, data=analytic2, family=poisson)
summ(fit1, confint = TRUE, digits = 3)

# Poisson regression - adjusted for covariates
fit2 <- glm(fruit.cont ~ phyact2 + age + sex + 
               income + race + bmicat + 
               smoke + edu, data=analytic2, family=poisson)
summ(fit2, confint = TRUE, digits = 3, vifs = TRUE)
```

### Survey weighted Poisson regression

Now, let's fit the design-adjusted Poisson regression using the `svyglm` function:

```{r prs, cache=TRUE}
require(jtools)
# Design
w.design <- update (w.design , phyact2=relevel(phyact, ref ="Inactive"))

# Design-adjusted Poisson - crude
fit1 <- svyglm(fruit.cont ~phyact2, design=w.design, family=poisson)
summ(fit1, confint = TRUE, digits = 3)

# Design-adjusted Poisson - adjusted for covariates
fit2<-svyglm(fruit.cont ~phyact2 + age + sex + 
               income + race + bmicat + 
               smoke + edu, design=w.design, family=poisson)
summ(fit2, confint = TRUE, digits = 3, vifs = TRUE)
```

### Negative binomial regression

Let's fit the negative binomial regression model using the `MASS::glm.nb` function, which estimates the dispersion parameter (theta, $\theta$) from the data rather than fixing it. In the negative binomial model the variance is $\text{Var}(Y) = \mu + \mu^2/\theta$, so the variance exceeds the mean whenever $\theta$ is finite, allowing the model to accommodate overdispersion. As $\theta \to \infty$ the variance approaches $\mu$ and the model reduces to Poisson. We report the estimated $\theta$ alongside the rate ratios.

```{r nb, cache=TRUE}
require(MASS)
analytic2$phyact2=relevel(analytic2$phyact, ref ="Inactive")

# Negative binomial regression - crude
fit3<- glm.nb(fruit.cont ~ phyact2, data=analytic2)
fit3$theta # estimated dispersion parameter (theta)
round(exp(cbind(coef(fit3), confint(fit3))),2)

# Negative binomial regression - adjusted for covariates
fit4<-glm.nb(fruit.cont ~phyact2 + age + sex + income + race + bmicat + smoke + edu, data=analytic2)
fit4$theta # estimated dispersion parameter (theta)
round(exp(cbind(coef(fit4), confint(fit4))),2)
```

The exponentiated coefficients are rate ratios (IRRs). For example, the `phyact2` rate ratio compares the expected fruit consumption count for the corresponding physical-activity level relative to the inactive reference group, holding other covariates fixed.

### Survey weighted negative binomial regression

Now, let's fit the design-adjusted negative binomial regression model. The `svyglm.nb` fit returns the estimated dispersion parameter (theta) together with the mean-model coefficients (prefixed with `eta.`). We report theta separately and exponentiate only the mean-model (`eta.`) coefficients to obtain rate ratios; the `theta.(Intercept)` row is not a rate ratio and should not be exponentiated.

```{r nbs, cache=TRUE}
require(sjstats)

# Helper: exponentiate only the mean-model (eta.) coefficients as rate ratios
nb_irr <- function(fit){
  est <- cbind(coef(fit), confint(fit))
  eta.rows <- grepl("^eta\\.", rownames(est))
  round(exp(est[eta.rows, , drop = FALSE]), 2)
}

# Design-adjusted negative binomial - crude
fit3<- svyglm.nb(fruit.cont ~phyact2, design=w.design)
attr(fit3, "nb.theta") # estimated dispersion parameter (theta)
nb_irr(fit3)

# Design-adjusted negative binomial - adjusted for covariates
fit4<-svyglm.nb(fruit.cont ~phyact2 + age + sex + income + race + bmicat + smoke + edu,
                design=w.design)
attr(fit4, "nb.theta") # estimated dispersion parameter (theta)
nb_irr(fit4)
```

As above, these exponentiated `eta.` coefficients are design-adjusted rate ratios (IRRs); a rate ratio above 1 indicates higher expected counts relative to the inactive reference group, and a value below 1 indicates lower expected counts.

### Video content (optional)

::: callout-tip
For those who prefer a video walkthrough, feel free to watch the video below, which offers a description of an earlier version of the above content.
:::

::: {style="position: relative; padding-bottom: 56.25%; height: 0; overflow: hidden;"}
<iframe src="https://www.youtube.com/embed/p2r8N7tcXU8" style="position: absolute; top: 0; left: 0; width: 100%; height: 100%; border:0;" allowfullscreen>

</iframe>
:::
