## Collinear predictors {.unnumbered}

```{r setup, warning=FALSE, cache=TRUE, message=FALSE}
# Load required packages
library(rms)
library(Hmisc)
```

### Load data

Let us load the dataset and see structure of the variables:

```{r load, warning=FALSE, cache=TRUE}
load(file = "Data/predictivefactors/cholesterolNHANES15.RData")
#head(analytic)
str(analytic)
```

### Describe the data

```{r data, warning=FALSE, cache=TRUE}
require(rms)
describe(analytic) 
```

### Identify collinear predictors

::: callout-tip
We can use `hclust` and `varclus` or variable clustering, i.e., to identify collinear predictors
:::

::: column-margin
`hclust` is the hierarchical clustering function where default is squared Spearman correlation coefficients to detect monotonic but nonlinear relationships.
:::

```{r collinear, warning=FALSE, cache=TRUE}
require(Hmisc)
sel.names <- c("gender", "age", "born", "race", "education", "married", 
               "income", "diastolicBP", "systolicBP", 
               "bodyweight", "bodyheight", "bmi", "waist", "smoke", "alcohol", 
               "cholesterol", "triglycerides", "uric.acid", 
               "protein", "bilirubin", "phosphorus", "sodium", "potassium", 
               "globulin", "calcium", "physical.work", "physical.recreational", 
               "diabetes")
var.cluster <- varclus(~., data = analytic[sel.names])
# var.cluster
plot(var.cluster)
```

### Video content (optional)

::: callout-tip
For those who prefer a video walkthrough, feel free to watch the video below, which offers a description of an earlier version of the above content.
:::

::: {style="position: relative; padding-bottom: 56.25%; height: 0; overflow: hidden;"}
<iframe src="https://www.youtube.com/embed/5nNZpsuQ_Fg" style="position: absolute; top: 0; left: 0; width: 100%; height: 100%; border:0;" allowfullscreen>

</iframe>
:::
