## Formatting Tools {.unnumbered}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DT)
```

The tutorial elucidates a variety of tools and methodologies aimed at streamlining and enhancing academic writing and presentation creation, discussing topics such as utilizing a typesetting system, converting tables across different formats, employing various R packages for enhanced data visualization and presentation, drawing flow and Gantt charts, crafting HTML5 presentations, enabling collaborative writing and document sharing, managing references efficiently, formatting articles, and employing specific platforms for identifying appropriate journals for publishing academic articles.

### LaTex

Get an account in ShareLaTeX/[Overleaf](https://v2.overleaf.com/login).

### Table conversion

From MS word to latex / markdown / HTML in [TablesGenerator](http://www.tablesgenerator.com/).

You can use `tableone` package to generate csv file, and then import them in the TablesGenerator to convert to HTML to paste to doc file!

```{r, eval=FALSE}
tab1x <- print(tab1, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
write.csv(tab1x, file = "tab1x.csv")
```

### Fancy table and chart generators:

R Packages

-   officer
-   officedown
-   flextable
-   mschart

### Drawing flow chart

[draw.io](https://www.draw.io/.)

### Presentation

The "xaringan" package, derived from a love for the Japanese manga and anime "Naruto", serves as an R Markdown extension, and it facilitates the creation of distinctively styled HTML5 presentations by leveraging the JavaScript library remark.js. Originating from an intent to produce a unique, though not widely adopted style, due to its potentially challenging pronunciation unless familiar with the anime, "xaringan" offers significant customizability in presentation design and has garnered additional theme contributions from its user community. Despite only supporting Markdown, "xaringan" enhances remark.js by introducing support for R Markdown and additional utilities, simplifying the slide-building and previewing processes. Further insights into "xaringan", its background, and its utility can be explored through its [documentation](http://slides.yihui.name/xaringan/).

### Gantt charts

[officetimeline](https://online.officetimeline.com/)

### Simultaneous collaborative writing

Google Docs offers a platform for real-time collaborative writing. Multiple users can edit documents simultaneously, and changes are saved automatically.

-   Commenting and Suggesting: Use the comment and suggest features to provide feedback without altering the original text.
-   Revision History: Navigate through the revision history to view changes and revert to previous versions if needed.
-   Sharing and Permissions: Manage who can view, comment, or edit the document with varied permission levels.

### Reference manager

[Zotero](https://www.zotero.org/) stands out as a free, open-source reference management software that assists researchers, academics, and students in organizing, managing, and formatting their citations and bibliographies. It's not just a reference manager but also a powerful tool for collaborative work on research projects. Try the [Zotero desktop manager](https://www.zotero.org/download/) as well for assisting with reference inserting.

Zotero syncs data across devices, ensuring that users can access their libraries from any location. Users can work offline with Zotero, and any changes made will be synchronized when the internet connection is restored.

[ZoteroBib](https://zbib.org/): Use ZoteroBib to generate bibliographies instantly without creating an account or installing software.

### Article formatting

The `rticles` package in R provides a diverse selection of templates for creating academic articles and is easily accessible directly within the RStudio environment by navigating through **File -\> New File -\> R Markdown**, where users can select their desired template. For users not utilizing RStudio, the installation of Pandoc is requisite, with articles being creatable using the `rmarkdown::draft()` function, and specifying the template and package parameters as needed. Additionally, the package enables viewing a list of available journal names using `rticles::journals()`. To employ enhanced features, such as automatic figure numbering and cross-referencing of tables, users can utilize functionalities from the `bookdown` package. This involves adjusting the YAML to use `bookdown::pdf_book` as the output format, and designating the chosen `rticles` template as the `base_format`. Comprehensive details and tutorials regarding the use of the `rticles` package can be found in its online documentation. The complete array of options can be explored within the R Markdown templates window in RStudio, via the packages's [GitHub readme](https://github.com/rstudio/rticles) or accessed programmatically via the following function:

```{r}
rticles::journals()
```

### Find appropriate journals

-   [jane](http://jane.biosemantics.org)
-   See more extensive list [here](https://ehsanx.github.io/Scientific-Writing-for-Health-Research/finding-suitable-journals.html)
-   Specific [Epidemiology-focus journals](https://ehsanx.github.io/choosing-epidemiological-journals/)

### Summary

| Category                     | Tool                                                                                                          | Description                                                                 |
|------------------------------|---------------------------------------------------------------------------------------------------------------|-----------------------------------------------------------------------------|
| Academic Search Engine       | [Google Scholar](https://scholar.google.com)                                                                  | Freely accessible search engine indexing scholarly literature.              |
| Article Formatting           | [rticles (R Package)](https://github.com/rstudio/rticles)                                                     | R package providing templates for various academic journals.                |
| Brainstorming & Mapping      | [MindMeister](https://www.mindmeister.com)                                                                    | Online mind mapping tool for brainstorming and collaborative visualization. |
| Document Creation & Editing  | [Overleaf](https://v2.overleaf.com/login)                                                                     | Collaborative LaTeX editor online.                                          |
|                              | [Google Docs](https://docs.google.com)                                                                        | Platform for real-time collaborative writing.                               |
|                              | [officer (R Package)](https://cran.r-project.org/web/packages/officer/readme/README.html)                     | R package for generating Word and PowerPoint files.                         |
|                              | [officedown (R Package)](https://cran.r-project.org/web/packages/officedown/index.html)                       | R package to produce Word documents with officer.                           |
| Gantt Chart Generators       | [Office Timeline](https://online.officetimeline.com/)                                                         | Online tool for creating Gantt charts.                                      |
| Journal Finding              | [jane](http://jane.biosemantics.org)                                                                          | Tool to assist in finding the right journal for publishing.                 |
|                              | [Custom List](https://ehsanx.github.io/Scientific-Writing-for-Health-Research/finding-suitable-journals.html) | Extensive list and guide on finding suitable journals.                      |
|                              | [Epidemiology Journals](https://ehsanx.github.io/choosing-epidemiological-journals/)                          | Specific list of epidemiology-focus journals.                               |
| Note & Research Management   | [Evernote](https://evernote.com)                                                                              | Note-taking and organization tool for managing research notes and drafts.   |
| Presentation & Sharing       | [Prezi](https://www.prezi.com)                                                                                | Dynamic and visually engaging presentation creation tool.                   |
|                              | [SlideShare](https://www.slideshare.net)                                                                      | Platform for sharing presentations and professional documents.              |
| Presentation                 | [xaringan (R Package)](http://slides.yihui.name/xaringan/)                                                    | R Markdown extension for creating presentations using remark.js.            |
| Project Management           | [Asana](https://asana.com)                                                                                    | Project management tool for workflow organization and collaboration.        |
|                              | [Trello](https://trello.com)                                                                                  | Project management tool for task tracking and collaboration.                |
| Reference Management         | [EndNote](https://endnote.com)                                                                                | Reference management software for organizing and integrating references.    |
|                              | [JabRef](https://www.jabref.org)                                                                              | Open-source bibliography reference manager using BibTeX.                    |
|                              | [Mendeley](https://www.mendeley.com)                                                                          | Reference management and academic social network.                           |
|                              | [Paperpile](https://paperpile.com)                                                                            | Reference management and academic research library.                         |
|                              | [Zotero](https://www.zotero.org/)                                                                             | Reference management and collaborative tool.                                |
|                              | [ZoteroBib](https://zbib.org/)                                                                                | Quick bibliography generation tool.                                         |
| Research Identity Management | [ORCID](https://orcid.org)                                                                                    | Provides a persistent digital identifier to distinguish researchers.        |
| Table & Chart Generators     | [draw.io](https://www.draw.io/)                                                                               | Online diagram software for flow charts and various diagrams.               |
|                              | [TablesGenerator](http://www.tablesgenerator.com/)                                                            | Converts tables to LaTeX, markdown, HTML formats.                           |
|                              | [flextable (R Package)](https://cran.r-project.org/web/packages/flextable/index.html)                         | R package for tabular reporting in various formats (Word, HTML, etc.).      |
|                              | [mschart (R Package)](https://cran.r-project.org/web/packages/mschart/index.html)                             | R package to create PowerPoint charts.                                      |
| Writing & Editing            | [Authorea](https://www.authorea.com)                                                                          | Collaborative platform for writing, citing, and publishing.                 |
|                              | [Grammarly](https://www.grammarly.com)                                                                        | Writing assistant for grammar and style enhancement.                        |
