# Research questions

## Background

When we are starting a research project, one of the first steps is to clearly define your research topic or question. According to Thabane et al., there is a structured approach to do this, primarily using the PICOT and FINER frameworks.

::: column-margin
@thabane2009posing
:::

1.  **PICOT Framework**:

The PICOT framework helps to structure a specific and clear research question by focusing on five key elements:

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
picot_data <- data.frame(
  Element = c("P", "I", "C", "O", "T"),
  Description = c(
    "Population of Interest: Who is the target group you are studying?",
    "Intervention: What is the main action, treatment, or variable you're looking at?",
    "Comparison: Are you comparing the intervention against a control group or usual care?",
    "Outcome of Interest: What specifically do you want to measure?",
    "Time Frame: Over what time period will your study take place?"
  ),
  Example = c(
    "US adults",
    "Effect of having rheumatoid arthritis",
    "People without rheumatoid arthritis",
    "Rate of cardiovascular diseases",
    "1999–2018"
  )
)

kable(picot_data, col.names = c("Element", "Description", "Example"))%>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

::: column-margin
Research Question: "In US adults, does having rheumatoid arthritis, compared to those without rheumatoid arthritis, affect the rate of cardiovascular diseases during 1999--2018?" based on @hossain2022association: [DOI: 10.1016/j.annepidem.2022.03.005](https://doi.org/10.1016/j.annepidem.2022.03.005)
:::

2.  **FINER Criteria**:

Once we have formulated your research question with the help of the PICOT elements, we should evaluate it using the FINER criteria:

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
finer_data <- data.frame(
  Element = c("F", "I", "N", "E", "R"),
  Description = c(
    "Feasible: Is it possible to conduct this research with available resources?",
    "Interesting: Is the research question intriguing to the scientific community?",
    "Novel: Is the question original and not already thoroughly researched?",
    "Ethical: Is the research ethically sound?",
    "Relevant: Is the research currently needed or will it fill a gap in existing knowledge?"
  )
)
kable(finer_data, col.names = c("Element", "Description"), caption = "FINER Criteria")%>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

The key takeaway is: Use the PICOT and FINER frameworks to guide you in framing a compelling, ethical, and achievable research question.

::: column-margin
Refer to the ['Scientific Writing for Health Research' book chapter](https://ehsanx.github.io/Scientific-Writing-for-Health-Research/research-question.html) for more details and examples.
:::

::: callout-note
We include 2 types of tutorials that emphasize the critical steps of **data preparation** and analysis tailored to specific research questions, cosidering the PICOT framework. They underscore the importance of refining and cleaning datasets to ensure their suitability for rigorous analytical procedures. The analyses, while rooted in distinct methodologies, converge on the common goal of deriving meaningful insights and ensuring the integrity and validity of the results obtained from the processed analytical data.
:::

::: column-margin
**Data preparation**: Merging, reformatting and recategorizing essential variables to create a dataset suitable for analysis, aligning it with the study's objectives.
:::

::: callout-important
**Datasets**:

All of the datasets used in this tutorial can be accessed from [this GitHub repository folder](https://github.com/ehsanx/EpiMethods/tree/main/Data/researchquestion)
:::

## Overview of tutorials

### Predictive questions

::: column-margin
In this chapter, we've embarked on a journey to understand the nuances of different research questions, laying the groundwork for the topics that lie ahead. As we move forward, the next chapter will delve deeper into the challenges associated with [causal questions](confounding.html). We'll explore the complexities of causal associations and discuss the optimal types of variables to include in adjustment models for accurate treatment effect estimation. Following that, we'll transition to a chapter dedicated entirely to [predictive](predictivefactors.html) questions, shedding light on their unique attributes and the methodologies best suited for addressing them. Join us as we navigate these intricate terrains of research inquiry.
:::

The [first tutorial](researchquestion1.html) serves to educate the user on how to utilize the RHC dataset to answer a **predictive** research question: developing a prediction model for the length of stay. The tutorial equips users with the skills to clean and process raw data, transforming it into an analyzable format, and introduces concepts that will be foundational for subsequent analysis.

The second tutorial ([part a for downloading](researchquestion2a.html) and [part b for analyzing](researchquestion2b.html)) provides an in-depth guide on how to build a **predictive** model for Diastolic blood pressure using the NHANES dataset for the years 2013-14.

### Causal questions

The [third tutorial](researchquestion3.html) aims to guide a study on the relationship between Osteoarthritis (OA) and cardiovascular diseases (CVD) among Canadian adults from 2001-2005. Utilizing the Canadian Community Health Survey (CCHS) cycle 1.1-3.1, the study intends to explore whether OA **increases** (more accurately, whether **associated** with) the risk of developing CVD.

The NHANES dataset was analyzed in this [forth tutorial](researchquestion4.html) to explore the relationship between health predictors and cholesterol levels (**association/causal**). After refining the survey design and handling missing data, regression models were built using varying predictors. Standard error computations and p-values were derived, adjusting for the survey's unique structure.

::: callout-tip
**Optional Content**:

You'll find that some sections conclude with an optional video walkthrough that demonstrates the code. Keep in mind that the content might have been updated since these videos were recorded. Watching these videos is optional.
:::

::: callout-warning
**Bug Report**:

Fill out [this form](https://forms.gle/YSwuiebtb1E9wjHu9) to report any issues with the tutorial.
:::

**Reference**
