## Concepts (Q) {.unnumbered}

According to Thabane et al., there is a structured approach to do this, primarily using the PICOT and FINER frameworks.

::: column-margin
@thabane2009posing
:::

## PICOT Framework

The PICOT framework helps to structure a specific and clear research question by focusing on five key elements:

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
picot_data <- data.frame(
  Element = c("P", "I", "C", "O", "T"),
  Description = c(
    "Population of Interest: Who is the target group you are studying?",
    "Intervention: What is the main action, treatment, or variable you're looking at?",
    "Comparison: Are you comparing the intervention against a control group or usual care?",
    "Outcome of Interest: What specifically do you want to measure?",
    "Time Frame: Over what time period will your study take place?"
  ),
  Example = c(
    "US adults",
    "Effect of having rheumatoid arthritis",
    "People without rheumatoid arthritis",
    "Rate of cardiovascular diseases",
    "1999–2018"
  )
)

kable(picot_data, col.names = c("Element", "Description", "Example"))%>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

::: column-margin
Research Question: "In US adults, does having rheumatoid arthritis, compared to those without rheumatoid arthritis, affect the rate of cardiovascular diseases during 1999--2018?" based on @hossain2022association: [DOI: 10.1016/j.annepidem.2022.03.005](https://doi.org/10.1016/j.annepidem.2022.03.005)
:::

## FINER Criteria

Once we have formulated your research question with the help of the PICOT elements, we should evaluate it using the FINER criteria:

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
finer_data <- data.frame(
  Element = c("F", "I", "N", "E", "R"),
  Description = c(
    "Feasible: Is it possible to conduct this research with available resources?",
    "Interesting: Is the research question intriguing to the scientific community?",
    "Novel: Is the question original and not already thoroughly researched?",
    "Ethical: Is the research ethically sound?",
    "Relevant: Is the research currently needed or will it fill a gap in existing knowledge?"
  )
)
kable(finer_data, col.names = c("Element", "Description"), caption = "FINER Criteria")%>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

The key takeaway is: Use the PICOT and FINER frameworks to guide you in framing a compelling, ethical, and achievable research question.

## SAP

A Statistical Analysis Plan (SAP), also referred to as a Data Analysis Plan (DAP) or Reporting Analysis Plan (RAP), is an integral part of research, particularly in randomized controlled trials (RCTs) [@kahan2020public], but also in observational studies [@hiemstra2019debate]. Here are a few reasons why it is beneficial to pre-plan the SAP for an observational study:

1.  Pre-planning an SAP helps define the specific analytical strategies and methods that will be used to answer the research questions. It outlines the techniques for handling data, including

-   the treatment of missing data, outliers,
-   the use of statistical tests, and
-   confounding adjustment techniques.

2.  By detailing the analysis plan before the data is examined, researchers ensure transparency and reduce the risk of data dredging or p-hacking.
3.  Confounding is a more pronounced issue in observational studies. Strategies for addressing confounding need to be more elaborate and explicit in observational studies.

::: column-margin
Refer to the ['Scientific Writing for Health Research' book chapter](https://ehsanx.github.io/Scientific-Writing-for-Health-Research/research-question.html) for more details and examples for PICOT, FINER and Statistical Analysis Plan (SAP).
:::

::: callout-note
We include 2 types of tutorials that emphasize the critical steps of **data preparation** and analysis tailored to specific research questions, cosidering the PICOT framework. They underscore the importance of refining and cleaning datasets to ensure their suitability for rigorous analytical procedures. The analyses, while rooted in distinct methodologies, converge on the common goal of deriving meaningful insights and ensuring the integrity and validity of the results obtained from the processed analytical data.
:::

::: column-margin
**Data preparation**: Merging, reformatting and recategorizing essential variables to create a dataset suitable for analysis, aligning it with the study's objectives.
:::

## Video Lessons

::: callout-tip
### PICOT and FINER

What is included in this Video Lesson:

-   References 0:53
-   How to get an idea about a Research Question? 1:05
-   Why the question need to be good? 2:41
-   A framework for defining a research question 5:17
-   Think hard about the ‘Outcome’ 14:40
-   Is this research doable? 17:57
-   Overall Roadmap 19:57
-   Other Reference (optional) 21:27

The timestamps are also included in the YouTube video description.
:::

::: {style="position: relative; padding-bottom: 56.25%; height: 0; overflow: hidden;"}
<iframe src="https://www.youtube.com/embed/eU6IyhwTVbk" style="position: absolute; top: 0; left: 0; width: 100%; height: 100%; border:0;" allowfullscreen>

</iframe>
:::

::: callout-tip
### SAP

What is included in this Video Lesson:

-   SAP 0:03
-   SAP example from a RCT 1:31
-   SAP example from an observational study 4:40
-   Code book 15:35

The timestamps are also included in the YouTube video description.
:::

::: {style="position: relative; padding-bottom: 56.25%; height: 0; overflow: hidden;"}
<iframe src="https://www.youtube.com/embed/F7DARSfZpgA" style="position: absolute; top: 0; left: 0; width: 100%; height: 100%; border:0;" allowfullscreen>

</iframe>
:::

## Video Lesson Slides

<iframe src="https://docs.google.com/presentation/d/e/2PACX-1vQ92qqeu7IC_EqN9Ma_crf8rtH0N97HOvTU66uIwP_TM8h5HJX3wTdQ07qK58281j225VNKbdcV6Oqf/embed?start=false&amp;loop=false&amp;delayms=3000" frameborder="0" width="672" height="398" allowfullscreen="true" mozallowfullscreen="true" webkitallowfullscreen="true">

</iframe>

## Links

-   [Google Slides](https://docs.google.com/presentation/d/1rZj_o6lBBqwlt4q26tNNXHSqyJMH7sdsHPkhFUG7tbo/edit?usp=sharing)
-   [PDF Slides](slides/researchQ.pdf)

## References
