## Model performance {.unnumbered}

```{r setup, warning=FALSE, message=FALSE, cache=TRUE}
# Load required packages
library(survey)
library(ROCR)
library(WeightedROC)
```

### Load data

```{r load, cache=FALSE}
load("Data/surveydata/cchs123w.RData")
load("Data/surveydata/cchs123w2.RData")
ls()
dim(analytic.miss)
dim(analytic2)
```

```{r fit0, cache=TRUE}
library(survey)
simple.model <- as.formula(I(CVD=="event") ~ OA)
fit0 <- svyglm(simple.model,
              design = w.design,
              family = binomial(logit))

basic.model
fit5 <- svyglm(basic.model,
              design = w.design,
              family = binomial(logit))

aic.int.model
fit9 <- svyglm(aic.int.model,
              design = w.design,
              family = binomial(logit))
```

### Model performance

#### ROC curve

-   The appropriateness of the fitted logistic regression model needs to be examined before it is accepted for use.
-   Plotting the pairs of - sensitivities vs - 1-specificities on a scatter plot provides a Receiver Operating Characteristic (ROC) curve.
-   The area under the ROC curve = AUC / C-statistic.
-   ROC/AUC should consider weights for complex surveys.

Grading Guidelines for AUC values:

-   0.90-1.0 excellent discrimination (unusual)
-   0.80-0.90 good discrimination
-   0.70-0.80 fair discrimination
-   0.60-0.70 poor discrimination
-   0.50-0.60 failed discrimination

```{r auc, cache=TRUE}
require(ROCR)
# WeightedROC may not be on cran for all R versions
# devtools::install_github("tdhock/WeightedROC")

library(WeightedROC)
svyROCw <- function(fit=fit,outcome=analytic2$CVD=="event", weight = NULL){
  # ROC curve for
  # Survey Data with Logistic Regression
  if (is.null(weight)){ # require(ROCR)
    prob <- predict(fit, type = "response")
  pred <- prediction(as.vector(prob), outcome)
  perf <- performance(pred, "tpr", "fpr")
  auc <- performance(pred, measure = "auc")
  auc <- auc@y.values[[1]]
  roc.data <- data.frame(fpr = unlist(perf@x.values), tpr = unlist(perf@y.values), 
      model = "Logistic")
  with(data = roc.data,plot(fpr, tpr, type="l", xlim=c(0,1), ylim=c(0,1), lwd=1,
     xlab="1 - specificity", ylab="Sensitivity",
     main = paste("AUC = ", round(auc,3))))
  mtext("Unweighted ROC")
  abline(0,1, lty=2)
  } else { # library(WeightedROC)
    outcome <- as.numeric(outcome)
  pred <- predict(fit, type = "response")
  tp.fp <- WeightedROC(pred, outcome, weight)
  auc <- WeightedAUC(tp.fp)
  with(data = tp.fp,plot(FPR, TPR, type="l", xlim=c(0,1), ylim=c(0,1), lwd=1,
     xlab="1 - specificity", ylab="Sensitivity",
     main = paste("AUC = ", round(auc,3))))
  abline(0,1, lty=2)
  mtext("Weighted ROC")
  }
}
```

```{r auc2, cache=TRUE}
summary(analytic2$weight)
analytic2$corrected.weight <- weights(w.design)
summary(analytic2$corrected.weight)
svyROCw(fit=fit0,outcome=analytic2$CVD=="event", weight = analytic2$corrected.weight)
svyROCw(fit=fit5,outcome=analytic2$CVD=="event", weight = analytic2$corrected.weight)
svyROCw(fit=fit9,outcome=analytic2$CVD=="event", weight = analytic2$corrected.weight)
# This function does not take in to account of strata/cluster
```

#### Archer and Lemeshow test

-   Hosmer Lemeshow-type tests are most useful as a very crude way to screen for fit problems, and should not be taken as a definitive diagnostic of a 'good' fit.
    -   problem in small sample size
    -   Dependent on G (group)
-   Archer and Lemeshow (2006) extended the standard Hosmer and Lemeshow GOF test for complex surveys.
-   After fitting the survey weighted logistic regression, the F-adjusted mean residual goodness-of-fit test could suggest
    -   no evidence of lack of fit (if P-value \> a reasonable cut-point, e.g., 0.05)
    -   evidence of lack of fit (if P-value \< a reasonable cut-point, e.g., 0.05)

```{r al1, cache=TRUE}
AL.gof <- function(fit=fit, data = analytic2, 
                   weight = "corrected.weight"){
  # Archer-Lemeshow Goodness of Fit Test for
  # Survey Data with Logistic Regression
  r <- residuals(fit, type="response") 
  f<-fitted(fit) 
  breaks.g <- c(-Inf, quantile(f,  (1:9)/10), Inf)
  breaks.g <- breaks.g + seq_along(breaks.g) * .Machine$double.eps
  g<- cut(f, breaks.g)
  data2g <- cbind(data,r,g)
  newdesign <- svydesign(id=~1, 
                         weights=as.formula(paste0("~",weight)), 
                        data=data2g)
  decilemodel<- svyglm(r~g, design=newdesign) 
  res <- regTermTest(decilemodel, ~g)
  return(res) 
}
```

```{r al2, cache=TRUE}
AL.gof(fit0, analytic2, weight ="corrected.weight")
AL.gof(fit5, analytic2, weight ="corrected.weight")
AL.gof(fit9, analytic2, weight = "corrected.weight")
```

#### Additional function

If the survey data contains strata and cluster, then the following function will be useful:

```{r gof2, cache=TRUE}
AL.gof2 <- function(fit=fit7, data = analytic, 
                   weight = "corrected.weight", psu = "psu", strata= "strata"){
  # Archer-Lemeshow Goodness of Fit Test for
  # Survey Data with Logistic Regression
  r <- residuals(fit, type="response") 
  f<-fitted(fit) 
  breaks.g <- c(-Inf, quantile(f,  (1:9)/10), Inf)
  breaks.g <- breaks.g + seq_along(breaks.g) * .Machine$double.eps
  g<- cut(f, breaks.g)
  data2g <- cbind(data,r,g)
  newdesign <- svydesign(id=as.formula(paste0("~",psu)),
                         strata=as.formula(paste0("~",strata)),
                         weights=as.formula(paste0("~",weight)), 
                        data=data2g, nest = TRUE)
  decilemodel<- svyglm(r~g, design=newdesign) 
  res <- regTermTest(decilemodel, ~g)
  return(res) 
}
```
