## Checking dataset {.unnumbered}

```{r setup, warning=FALSE, message=FALSE, cache=TRUE}
# Load required packages
library(survey)
library(knitr)
library(car)
library(tableone)
library(DataExplorer)
```

### Load data

```{r load, warning=FALSE, message=FALSE, cache = TRUE}
load("Data/surveydata/cchs123.RData")
ls()
```

### Checking

#### Check the data for missingness

```{r miss0, warning=FALSE, message=FALSE, cache = TRUE}
dim(analytic)
require("tableone")
#CreateTableOne(data = analytic, includeNA = TRUE)
CreateTableOne(data = analytic, strata = "CVD", includeNA = TRUE)
CreateTableOne(data = analytic, strata = "OA", includeNA = TRUE)

require(DataExplorer)
plot_missing(analytic)
```

#### Look for zero-cells

```{r cells0, warning=FALSE, message=FALSE, cache = TRUE}
analytic$age.65p <- analytic$age.teen <- 0
analytic$age.teen[analytic$age == "teen"] <- 1
analytic$age.65p[analytic$age == "65 years and over"] <- 1
CreateTableOne(data = analytic, strata = "age.teen", includeNA = TRUE)
# copd has zero cells
# analytic$age[analytic$age == 'teen'] <- NA (will set this if we use copd)
```

```{r tab0, warning=FALSE, message=FALSE, cache = TRUE}
CreateTableOne(data = analytic, strata = "age.65p", includeNA = TRUE)
analytic$age.65p <- analytic$age.teen <- NULL
```

```{r tab1, warning=FALSE, message=FALSE, cache = TRUE}
table(analytic$province.check,analytic$fruit)
table(analytic$age)
table(analytic$copd, analytic$age)
table(analytic$stress, analytic$age) 
```

-   universe 15 + is not an issue for `stress` as age starts from 20
-   `copd` is problematic!
-   But wasn't `fruit` variable measured in an optional component?

```{r tab2, warning=FALSE, message=FALSE, cache = TRUE}
table(analytic$province.check[analytic$cycle==11],
      analytic$fruit[analytic$cycle==11])
```

```{r tab3, warning=FALSE, message=FALSE, cache = TRUE}
table(analytic$province.check[analytic$cycle==21],
      analytic$fruit[analytic$cycle==21])
# a different QUEBEC spelling used
```

```{r tab4, warning=FALSE, message=FALSE, cache = TRUE}
table(analytic$province.check[analytic$cycle==31],
      analytic$fruit[analytic$cycle==31])
# The real problem!
```

-   Look at data dictionaries in all cycles
    -   cycle 1.1 FVCADTOT Universe: All respondents
    -   cycle 2.1 FVCCDTOT Universe: All respondents
    -   cycle 3.1 FVCEDTOT Universe: Respondents with FVCEFOPT = 1

Need to make a decision about what to do about this. Delete or keep/impute later?

```{r analytic0, warning=FALSE, message=FALSE, cache = TRUE}
dim(analytic)
analytic1 <- analytic
# analytic1$South[analytic1$province.check == "NFLD & LAB."] <- NA
# analytic1$South[analytic1$province.check == "YUKON/NWT/NUNA."] <- NA
# analytic1 <- subset(analytic, province.check != "NFLD & LAB." & 
#                       province.check != "YUKON/NWT/NUNA." )
dim(analytic1)

analytic1$copd <- NULL # will bring this later for missing data analysis
# CreateTableOne(data = analytic1, strata = "OA", includeNA = TRUE)
# analytic1 <- droplevels.data.frame(analytic1)
analytic1$province.check <- NULL # we already have simplified province variable
# CreateTableOne(data = analytic1, strata = "OA", includeNA = TRUE)
```

#### Set appropriate reference

Save the original data (with missing values)!

```{r analytic2, warning=FALSE, message=FALSE, cache = TRUE}
analytic.miss <- analytic1
```

```{r analytic3, warning=FALSE, message=FALSE, cache = TRUE}
analytic.miss$smoke <- relevel(as.factor(analytic.miss$smoke), ref='Never smoker')
analytic.miss$drink <- relevel(as.factor(analytic.miss$drink), ref='Never drank')
analytic.miss$province <- relevel(as.factor(analytic.miss$province), ref='South')
analytic.miss$immigrate <- relevel(as.factor(analytic.miss$immigrate), ref='not immigrant')
```

#### Complete data options

```{r analytic4, warning=FALSE, message=FALSE, cache = TRUE}
# Wrong thing to do for survey data analysis!!
analytic2 <- as.data.frame(na.omit(analytic1)) 
dim(analytic2)
# tab1 <- CreateTableOne(data = analytic2, strata = "OA", includeNA = TRUE)
# print(tab1, test=FALSE, showAllLevels = TRUE)
```

### Saving dataset

```{r saveall, warning=FALSE, message=FALSE, cache = TRUE}
dim(cc123a)
dim(analytic)
dim(analytic.miss)
dim(analytic2)
save(analytic.miss, analytic2, file = "Data/surveydata/cchs123b.RData")
```
