## CCHS: Bivariate analysis {.unnumbered}

The following tutorial is performing bivariate analysis on our CCHS analytic dataset to examine relationships between two variables (association question). 

We load several R packages required for bivariate analysis, statistical tests, and data visualization.

```{r setup, warning=FALSE, message=FALSE, cache=TRUE}
# Load required packages
library(survey)
library(knitr)
library(car)
library(tableone)
library(DataExplorer)
library(Publish)
library(ROCR)
library(WeightedROC)
library(jtools)
```

### Load data

We load the dataset into the R environment and lists all available variables and objects. 

```{r load, warning=FALSE, message=FALSE, cache = TRUE}
load("Data/surveydata/cchs123b.RData")
ls()
```

### Preparing data

#### Weights

Here, the weights of survey respondents are accumulated, to account for the combination of different cycles of the data.

```{r wgt, warning=FALSE, message=FALSE, cache = TRUE}
analytic.miss$weight <- analytic.miss$weight/3 # 3 cycles combined
```

#### Fixing variable types

We convert several variables to categorical or "factor" types, which are better suited for some statistical analysis when variables have categories.

```{r vartype, warning=FALSE, message=FALSE, cache = TRUE}
var.names <- c("CVD", "age", "sex", "married", "race", "edu", "income", "bmi", 
               "phyact", "doctor", "stress", "smoke", "drink", "fruit", "bp", 
               "diab", "province", "OA", "immigrate")
analytic.miss[var.names] <- lapply(analytic.miss[var.names] , factor)
str(analytic.miss)
```

The code identifies rows where data is missing and labels them for later analyses.

```{r vartype2, warning=FALSE, message=FALSE, cache = TRUE}
analytic.miss$miss <- 1
head(analytic.miss$ID) # full data
head(analytic2$ID) # complete case
head(analytic.miss$ID[analytic.miss$ID %in% analytic2$ID])
analytic.miss$miss[analytic.miss$ID %in% analytic2$ID] <- 0
table(analytic.miss$miss)
```

### Setting Design

The code sets up the survey design, specifying weights (but no specific clustering and stratification, as they are unavailable for CCHS public access data), for use in survey-weighted analyses.

```{r desgin0, warning=FALSE, message=FALSE, cache = TRUE}
require(survey)
summary(analytic.miss$weight)
w.design0 <- svydesign(id=~1, weights=~weight, 
                      data=analytic.miss)
summary(weights(w.design0))
sd(weights(w.design0))
```

This creates a subset of the data where there are no missing values. Note that subset was done to the design object `w.design0`, not the data `analytic.miss`.

```{r desgin1, warning=FALSE, message=FALSE, cache = TRUE}
w.design <- subset(w.design0, miss == 0)
summary(weights(w.design))
sd(weights(w.design))
```

### Bivariate analysis

#### Table 1 (weighted)

##### Stratified by exposure

These tables contain descriptive statistics, stratified by different categories. They can be useful for understanding how variables relate to the exposure or outcome in the data.

```{r tab2, warning=FALSE, message=FALSE, cache = TRUE}
require(tableone)
var.names <- c("CVD", "age", "sex", "married", "race", "edu", "income", "bmi", 
               "phyact", "doctor", "stress", "smoke", "drink", "fruit", "bp", 
               "diab", "province", "immigrate") # exclude "OA"
# tab1 <- CreateTableOne(var = var.names, strata= "OA", data=analytic.miss, test = TRUE)
# print(tab1)
tab2 <- svyCreateTableOne(var = var.names, strata= "OA", 
                          data=w.design, test = TRUE)
print(tab2)
```

##### Stratified by outcome

This table is generally useful for logistic regression analysis

```{r tab3, warning=FALSE, message=FALSE, cache = TRUE}
var.names <- c("OA", "age", "sex", "married", "race", "edu", "income", "bmi", 
               "phyact", "doctor", "stress", "smoke", "drink", "fruit", "bp", 
               "diab", "province", "immigrate") # exclude "CVD"
tab3 <- svyCreateTableOne(var = var.names, strata= "CVD", data=w.design, test = TRUE)
print(tab3)
```

How did they calculate the p-values? Hint: `svychisq` (see below).

#### Proportions and Design Effect

This part computes proportions and design effects, which help understand the influence of the sampling design on the estimated statistics.

```{r fit0, warning=FALSE, message=FALSE, cache = TRUE}
require(survey)
# Computing survey statistics on subsets of a survey defined by factor(s).
fit0a <- svyby(~CVD,~OA,design=w.design, svymean,deff=TRUE)
fit0a
confint(fit0a)
# 7.45% OA patients estimated to have CVD event.
# 95% CI:  (0.067, 0.0816)
```

Let

-   $\theta$ = parameter (population slope) and\
-   $\hat(\theta)$ = statistic (estimated slope).

$b = \frac{\sum[w (y_i-\bar{y}) (x_i-\bar{x})]}{\sum[w (x_i-\bar{x})^2]}$

DE = Effect of complex survey on the SEs, relative to a SRS of equal size.

-   $D^2(\hat{\theta}) = \frac{Var(\hat{\theta})_{Complex Survey}}{Var(\hat{\theta})_{SRS}}$
-   $D^2(\hat{\theta}) = \frac{SE(\hat{\theta})^2_{Complex Survey}}{SE(\hat{\theta})^2_{SRS}}$

Note:

1.  SE increases as value of weight increases (CCHS).
2.  NHANES has more things to worry about (strata, PSU)

DEFF = 2 means that the variance of the sample proportion, when choosing the sample by complex survey sampling, is nearly 2 times as large as the variance of the same estimator under simple random sampling/SRS.

```{r fit1, warning=FALSE, message=FALSE, cache = TRUE}
fit0b <- svyby(~CVD,~diab,design=w.design, svymean,deff=TRUE)
fit0b
confint(fit0b)
```

#### Testing association

Here, Chi-square tests are conducted to test the association between different variables. Two variants of the test are used: Rao-Scott and Thomas-Rao modifications. These adaptations are used when the data come from a complex survey design.

-   Tests for hypothesis
    -   Rao-Scott modifications (chi-sq)
    -   Thomas-Rao modifications (F)

```{r testchisq, warning=FALSE, message=FALSE, cache = TRUE}
# Rao-Scott modifications (chi-sq)
svychisq(~CVD+OA,design=w.design, statistic="Chisq")

# Thomas-Rao modifications (F)
svychisq(~CVD+OA,design=w.design, statistic="F") 

# Both provide strong evidence to reject the null hypothesis.
# Conclusion: there is a significant (at 5%) association 
# between CVD prevalence and OA.
svychisq(~CVD+fruit,design=w.design, statistic="F") 
svychisq(~CVD+province,design=w.design, statistic="Chisq") 
```

### Saving data

Finally, the dataset, along with any new variables or subsets created during the analysis, is saved for future use.

```{r save, cache = TRUE}
save(w.design, analytic.miss, analytic2, file = "Data/surveydata/cchs123w.RData")
```


### Video content (optional)

::: callout-tip
For those who prefer a video walkthrough, feel free to watch the video below, which offers a description of an earlier version of the above content.
:::

::: {style="position: relative; padding-bottom: 56.25%; height: 0; overflow: hidden;"}
<iframe src="https://www.youtube.com/embed/5q_GyL5FsVg" style="position: absolute; top: 0; left: 0; width: 100%; height: 100%; border:0;" allowfullscreen>

</iframe>
:::