## NHANES: Performance {.unnumbered}

This tutorial outlines how to evaluate the performance of a logistic regression model fitted to complex survey data using R, focusing on the NHANES dataset.

We will assess the model's predictive accuracy using a design-correct **Area Under the Curve (AUC)** and evaluate its fit with the **Archer-Lemeshow** test.

------------------------------------------------------------------------

## 1. Data Preparation ⚙️

First, we need to load the necessary R packages and prepare the NHANES data. The code below loads data from the `Flegal2016.RData` file, recodes variables like age, race, smoking, and education, and creates the final survey design object (`analytic_design`) for our analysis.

```{r, warning=FALSE, message=FALSE, cache=TRUE}
# --- Load Libraries ---
library(survey)
library(dplyr)
# install.packages("devtools")
# devtools::install_github("ehsanx/svyTable1", 
#                          build_vignettes = TRUE, 
#                          dependencies = TRUE)
library(svyTable1)
library(ggplot2)

# --- Load Data ---
load("Data/surveydata/Flegal2016.RData") 

# --- Create Analytic Sample (N=5,455) ---
dat.analytic <- subset(dat.full, RIDAGEYR >= 20)
dat.analytic <- subset(dat.analytic, !is.na(BMXBMI))
dat.analytic <- subset(dat.analytic, is.na(RIDEXPRG) | RIDEXPRG != "Yes, positive lab pregnancy test")

# --- Recode Analysis Variables ---
dat.full <- dat.full %>%
  mutate(
    Age = cut(RIDAGEYR,
              breaks = c(20, 40, 60, Inf),
              right = FALSE,
              labels = c("20-39", "40-59", ">=60")),

    race = case_when(
      RIDRETH3 == "Non-Hispanic White" ~ "Non-Hispanic white",
      RIDRETH3 == "Non-Hispanic Black" ~ "Non-Hispanic black",
      RIDRETH3 == "Non-Hispanic Asian" ~ "Non-Hispanic Asian",
      RIDRETH3 %in% c("Mexican American", "Other Hispanic") ~ "Hispanic",
      TRUE ~ "Other" # Consolidating "Other Race" for simplicity
    ),
    race = factor(race, levels = c("Non-Hispanic white", "Non-Hispanic black", "Non-Hispanic Asian", "Hispanic", "Other")),

    obese = factor(ifelse(BMXBMI >= 30, "Yes", "No"), levels = c("No", "Yes")),

    smoking = case_when(
      SMQ020 == "No" ~ "Never smoker",
      SMQ020 == "Yes" & SMQ040 == "Not at all" ~ "Former smoker",
      SMQ020 == "Yes" ~ "Current smoker",
      TRUE ~ NA_character_
    ),
    smoking = factor(smoking, levels = c("Never smoker", "Former smoker", "Current smoker")),

    education = case_when(
        DMDEDUC2 %in% c("Less than 9th grade", "9-11th grade (Includes 12th grad)") ~ "<High school",
        DMDEDUC2 == "High school graduate/GED or equi" ~ "High school",
        DMDEDUC2 %in% c("Some college or AA degree", "College graduate or above") ~ ">High school",
        TRUE ~ NA_character_
    ),
    education = factor(education, levels = c("High school", "<High school", ">High school"))
  )

# --- Create Survey Design Object ---
options(survey.lonely.psu = "adjust")
svy.design.full <- svydesign(id = ~SDMVPSU,
                             strata = ~SDMVSTRA,
                             weights = ~WTINT2YR,
                             nest = TRUE,
                             data = dat.full)
analytic_design <- subset(svy.design.full, SEQN %in% dat.analytic$SEQN)
```

------------------------------------------------------------------------

## 2. Fitting a Logistic Regression Model 📊

Next, we fit a survey-weighted logistic regression model using `svyglm()`. Our goal is to predict obesity status based on **Age**, **race**, **smoking status**, and **education level**.

```{r, cache=TRUE}
# Fit the survey-weighted logistic regression model
fit_obesity <- svyglm(
  I(obese == "Yes") ~ Age + race + smoking + education,
  design = analytic_design,
  family = quasibinomial()
)
```

------------------------------------------------------------------------

## 3. Model Performance: ROC Curve and AUC 📈

We calculate the **Area Under the ROC Curve (AUC)** to measure predictive performance. The `svyAUC()` function from `svyTable1` computes a design-corrected AUC and confidence interval, which requires a replicate-weights survey design.

### Create Replicate Design & Refit Model

```{r, cache=TRUE}
# svyAUC() requires a replicate-weights design object
rep_design <- as.svrepdesign(analytic_design)
# Restrict the replicate design to complete cases
vars_for_model <- c("obese", "Age", "race", "smoking", "education")
rep_design_complete <- subset(rep_design, 
                              complete.cases(rep_design$variables[, vars_for_model]))

# Refit model on the complete-case replicate design
fit_obesity_rep <- svyglm(
  I(obese == "Yes") ~ Age + race + smoking + education,
  design = rep_design_complete,
  family = quasibinomial()
)
```

### Calculate AUC

```{r, cache=TRUE}
# Calculate the design-correct AUC
auc_results_list <- svyAUC(fit_obesity_rep, rep_design_complete, plot = TRUE)

# Display the AUC summary
knitr::kable(auc_results_list$summary)
```

**Interpretation:** An AUC of **0.5967987** indicates poor discrimination. While better than simpler models, it suggests that the predictors do not have strong predictive power for obesity in this dataset.

### Visualize the ROC Curve

```{r, cache=TRUE}
ggplot(auc_results_list$roc_data, aes(x = FPR, y = TPR)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_abline(linetype = "dashed") +
  labs(
    title = "Survey-Weighted ROC Curve",
    subtitle = paste0("AUC = ", round(auc_results_list$summary$AUC, 3)),
    x = "1 - Specificity (FPR)",
    y = "Sensitivity (TPR)"
  ) +
  theme_minimal()
```

------------------------------------------------------------------------

## 4. Goodness-of-Fit: Archer and Lemeshow Test ✅

We assess model fit using the **Archer-Lemeshow test**. A non-significant p-value (p \> 0.05) suggests a good model fit.

```{r, cache=TRUE}
# Run the goodness-of-fit test
gof_results <- svygof(fit_obesity, analytic_design)

# Display the results
knitr::kable(gof_results)
```

**Interpretation:** The p-value (**0.1791469**) is greater than 0.05, indicating no evidence of a poor fit. The model’s predictions align well with observed outcomes, suggesting good calibration.

------------------------------------------------------------------------

## Summary

-   The AUC shows poor discriminatory power, meaning the model does not strongly distinguish between obese and non-obese individuals.
-   The Archer-Lemeshow test indicates good model fit (p \> 0.05).

**Conclusion:** While well-calibrated, the model’s low predictive power implies other factors (e.g., cholesterol, blood pressure, diet) may better explain obesity risk.
