## NHANES: Performance {.unnumbered}

This tutorial outlines how to evaluate the performance of a logistic regression model fitted to complex survey data using R, focusing on the NHANES dataset.

We will assess the model's predictive accuracy using a design-correct **Area Under the Curve (AUC)** and evaluate its fit with the **Archer-Lemeshow** test.

------------------------------------------------------------------------

## 1. Data Preparation ⚙️

First, we need to load the necessary R packages and prepare the NHANES data.

We'll use data for adults (age 20+) and create a binary outcome variable for obesity status.

### Load Packages

We'll need `svyTable1`, `survey`, and `dplyr` for data manipulation, and `NHANES` for the dataset.

```{r, Warnings=FALSE, message=FALSE}
library(svyTable1)
library(survey)
library(dplyr)
library(NHANES)
```

### Prepare Data

We load the raw NHANES data, filter for adults, and create an `ObeseStatus` factor.

For this modeling example, we'll create a "complete case" dataset by removing rows with missing values for simplicity.

```{r, cache=TRUE}
# Load the raw NHANES data (2009-2012)
data(NHANESraw)

# Prepare data for adults
nhanes_adults_with_na <- NHANESraw %>%
  dplyr::filter(Age >= 20) %>%
  mutate(
    ObeseStatus = factor(ifelse(BMI >= 30, "Obese", "Not Obese"),
                         levels = c("Not Obese", "Obese"))
  )

# Define variables for a complete-case analysis
vars_for_complete_table <- c("Age", "Gender", "Race1", "ObeseStatus")

# Create complete-case data frame
nhanes_adults_complete <- nhanes_adults_with_na %>%
  tidyr::drop_na(all_of(vars_for_complete_table))
```

### Create Survey Design Object

We now create a `svydesign` object, which accounts for the complex survey design (PSU, strata, and weights).

This step ensures accurate variance estimation and population inference.

```{r, cache=TRUE}
# Create a survey design object
adult_design_complete <- svydesign(
  id = ~SDMVPSU,
  strata = ~SDMVSTRA,
  weights = ~WTMEC2YR,
  nest = TRUE,
  data = nhanes_adults_complete
)
```

------------------------------------------------------------------------

## 2. Fitting a Logistic Regression Model 📊

Next, we fit a **survey-weighted logistic regression** model using `svyglm()`.

Our goal is to predict obesity status (`ObeseStatus`) based on **Age**, **Gender**, and **Race1**.

```{r, cache=TRUE}
# Fit the survey-weighted logistic regression model
fit_obesity <- svyglm(
  ObeseStatus ~ Age + Gender + Race1,
  design = adult_design_complete,
  family = quasibinomial()
)
```

------------------------------------------------------------------------

## 3. Model Performance: ROC Curve and AUC 📈

We calculate the **Area Under the ROC Curve (AUC)** to measure predictive performance.

The `svyAUC()` function from `svyTable1` computes a *design-corrected* AUC and confidence interval.

### Create Replicate Design & Refit Model

```{r, cache=TRUE}
# svyAUC() requires a replicate-weights design object
rep_design <- as.svrepdesign(adult_design_complete)

# Refit the model using replicate design
fit_obesity_rep <- svyglm(
  ObeseStatus ~ Age + Gender + Race1,
  design = rep_design,
  family = quasibinomial()
)
```

### Calculate AUC

```{r, cache=TRUE}
# Calculate design-correct AUC
auc_results_list <- svyAUC(fit_obesity_rep, rep_design, plot = TRUE)

# Display AUC summary
knitr::kable(auc_results_list$summary)
```

**Interpretation:**
An AUC of **0.587** indicates weak discrimination — the model performs only slightly better than random chance (AUC = 0.5).

This suggests that *Age*, *Gender*, and *Race1* alone are insufficient to accurately predict obesity.

### Visualize the ROC Curve

```{r, cache=TRUE}
library(ggplot2)

ggplot(auc_results_list$roc_data, aes(x = FPR, y = TPR)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_abline(linetype = "dashed") +
  labs(
    title = "Survey-Weighted ROC Curve",
    subtitle = paste0("AUC = ", auc_results_list$summary$AUC),
    x = "1 - Specificity (FPR)",
    y = "Sensitivity (TPR)"
  ) +
  theme_minimal()
```

------------------------------------------------------------------------

## 4. Goodness-of-Fit: Archer and Lemeshow Test ✅

We now assess how well the model's predictions match observed data using the **Archer-Lemeshow test**, an extension of the traditional Hosmer-Lemeshow test for complex surveys.

A non-significant p-value (p \> 0.05) suggests a good model fit.

```{r, cache=TRUE}
# Run the goodness-of-fit test
gof_results <- svygof(fit_obesity, adult_design_complete)

# Display the results
knitr::kable(gof_results)
```



**Interpretation:**
The p-value (0.0012) is **less than 0.05**, indicating that the model does **not** fit the data well. 

This suggests that it may be missing important predictors or interactions that explain obesity risk more accurately.

------------------------------------------------------------------------

### Summary

-   The **AUC** shows weak discriminatory power (≈ 0.59).
-   The **Archer-Lemeshow** test indicates **poor model fit** (p \< 0.05).
-   To improve, consider adding predictors such as *education*, *income*, or *smoking* and testing interaction effects.
