## CCHS: Revisiting PICOT {.unnumbered}

Welcome to this tutorial where we will examine the same research question presented in the [Causal question-1](researchquestion3.html) tutorial. Our approach will be enriched this time by working with a more comprehensive set of covariates. We will follow the guidelines from the research article by @rahman2013relationship ([DOI:10.1136/bmjopen-2013-002624](https://bmjopen.bmj.com/content/3/5/e002624)). We will also will work properly with survey feature variables (e.g., sampling weights).

::: column-margin
@rahman2013relationship
:::

### Remembering PICOT

Before diving into the data, let's clarify and remember the research parameters using the PICOT framework:

(1) **Target population**: Canadian adults (CCHS data)

(2) **Outcome** ($Y$): CVD (Heart disease/Cardiovascular Disease)

(3) **Exposure group** ($A$): Osteoarthritis (OA)

(4) **Control group**: People without OA

(5) **Timeline**: Data collected from 2001 to 2005

In addition, we'll identify potential confounders based on literature to better understand the relationship between exposure and outcome.

### Creating dataset {.unnumbered}

To start, we'll load the required R packages:

```{r setup, warning=FALSE, message=FALSE, cache=TRUE}
# Load required packages
library(survey)
library(knitr)
library(car)
```

### Load data

We'll be using CCHS data from various cycles. Use the following code to load this data into your R environment:

```{r data, warning=FALSE, message=FALSE, cache = FALSE}
# CCHS 1.1
load("Data/surveydata/cchsc1.RData")

# CCHS 2.1
load("Data/surveydata/cchsc2.RData")

# CCHS 3.1
load("Data/surveydata/cchsc3.RData")

# objects loaded
ls()
```

To check the dimensions of each data set:

```{r dim, warning=FALSE, message=FALSE, cache = TRUE}
# Dimensions
dim(c1)
dim(c2)
dim(c3)
```

### Subset the data (subset variables)

#### Understand variable coding

Before subsetting the data, we need to comprehend the variables used in CCHS 3.1, which are described in detail in the official reference:

-   page 60 in CCHS 3.1 guide [@statistics2005canadian]

-   link for variable list: [@varlist]

A table mapping variable concepts across different CCHS cycles can be found below.

| **Variable Concept**                                              | **CCHS 1.1** | **CCHS 2.1** | **CCHS 3.1** |
|---------------------|-----------------|-----------------|-----------------|
| **Has heart disease**                                             | CCCA_121     | CCCC_121     | CCCE_121     |
| **Has arthritis or rheumatism**                                   | CCCA_051     | CCCC_051     | CCCE_051     |
| **Kind of arthritis**                                             | CCCA_05A     | CCCC_05A     | CCCE_05A     |
| **Age**                                                           | DHHAGAGE     | DHHCGAGE     | DHHEGAGE     |
| **Sex**                                                           | DHHA_SEX     | DHHC_SEX     | DHHE_SEX     |
| **Marital Status**                                                | DHHAGMS      | DHHCGMS      | DHHEGMS      |
| **Cultural / racial origin**                                      | SDCAGRAC     | SDCCGRAC     | SDCEGCGT     |
| **Immigrant status**                                              | SDCAFIMM     | SDCCFIMM     | SDCEFIMM     |
| **Length of time in Canada since immigration**                    | SDCAGRES     | SDCCGRES     | SDCEGRES     |
| **Highest level of education - respondent**                       | EDUADR04     | EDUCDR04     | EDUEDR04     |
| **Total household income from all sources**                       | INCAGHH      | INCCGHH      | INCEGHH      |
| **Body mass index**                                               | HWTAGBMI     | HWTCGBMI     | HWTEGBMI     |
| **Physical activity index**                                       | PACADPAI     | PACCDPAI     | PACEDPAI     |
| **Has a regular medical doctor**                                  | TWDA_5       | HCUC_1AA     | HCUE_1AA     |
| **Self-perceived stress**                                         | GENA_07      | GENC_07      | GENE_07      |
| **Type of smoker**                                                | SMKADSTY     | SMKCDSTY     | SMKEDSTY     |
| **Type of drinker**                                               | ALCADTYP     | ALCCDTYP     | ALCEDTYP     |
| **Daily consumption - total fruits and vegetables**               | FVCADTOT     | FVCCDTOT     | FVCEDTOT     |
| **Has high blood pressure**                                       | CCCA_071     | CCCC_071     | CCCE_071     |
| **Has emphysema or chronic obstructive pulmonary disease (COPD)** | CCCA_91B     | CCCC_91B     | CCCE_91F     |
| **Has diabetes**                                                  | CCCA_101     | CCCC_101     | CCCE_101     |
| **Province**                                                      | GEOAGPRV     | GEOCGPRV     | GEOEGPRV     |
| **Sampling weight - master weight**                               | WTSAM        | WTSC_M       | WTSE_M       |

While most variables in CCHS 3.1 are universally applicable to 'All respondents,' there are some exceptions. For example:

-   CCCE_05A universe: (Kind of arthritis / rheumatism)

-   Respondents who answered CCCE_051 = (1, 7 or 8) or CCCE_011 = 8

-   CCCE_051: All respondents

-   CCCE_011: All respondents

-   SDCEGRES universe: (Length of time in Canada since immigration)

    -   Respondents who answered SDCE_2 = (2, 7 or 8) or SDCE_1 = (97 or 98)
    -   SDCE_2 doesn't exist!
    -   (master file variable; not available in PUMF)
    -   Public Use Microdata File (PUMF)
    -   SDCE_1 doesn't exist!
    -   (master file variable; not available in PUMF)

-   HWTEGBMI universe: (Body Mass Index (BMI) / self-report)

    -   All respondents excluding pregnant women (MAME_037 = 1)
        -   MAME_037 doesn't exist!
        -   (master file variable; not available in PUMF)

-   GENE_07 universe: (Self-perceived stress)

    -   Respondents aged 15 and over

-   CCCE_91F universe: (Has chronic obstructive pulmonary disease)

    -   Respondents aged 30 and over

-   FVCEDTOT universe: (Daily consumption - total fruits and vegetables)

    -   Respondents with FVCEFOPT = 1
        -   FVCEFOPT: Optional module: Fruit and vegetable consumption - (F)

Ref:

-   page 66 in CCHS 3.1 guide [@statistics2005canadian]

-   Potential problematic variables:

    -   Self-perceived stress
    -   Has chronic obstructive pulmonary disease / copd
    -   Daily consumption - total fruits and vegetables

We will make decisions about these variables later: for now, let's keep them.

#### Restrict the dataset with variables of interest only

##### Cycle 1.1

-   We define a vector of variable names `var.names1` that are of interest for the first cycle of the Canadian Community Health Survey (CCHS 1.1). These variables cover a range of topics such as heart disease, age, sex, etc.
-   Then we creates a new data frame `cc11` by subsetting the original data frame `c1` to include only the columns specified in `var.names1`.

```{r data1, warning=FALSE, message=FALSE, cache = TRUE}
var.names1 <- c("CCCA_121", "CCCA_051", "CCCA_05A", "DHHAGAGE", 
                "DHHA_SEX", "DHHAGMS", "SDCAGRAC", "SDCAFIMM", 
                "SDCAGRES", "EDUADR04", "INCAGHH", "HWTAGBMI", 
                "PACADPAI", "TWDA_5", "GENA_07", "SMKADSTY", 
                "ALCADTYP", "FVCADTOT", "CCCA_071", "CCCA_91B",
                "CCCA_101", "GEOAGPRV", "WTSAM")
cc11 <- c1[var.names1]
dim(cc11)
table(cc11$CCCA_051)
```

The subsequent two code chunks do the same for CCHS 2.1 and CCHS 3.1, respectively, resulting in new data frames `cc21` and `cc31`.

##### Cycle 2.1

```{r data2, warning=FALSE, message=FALSE, cache = TRUE}
var.names2 <- c("CCCC_121", "CCCC_051", "CCCC_05A", "DHHCGAGE", 
                "DHHC_SEX", "DHHCGMS", "SDCCGRAC", "SDCCFIMM", 
                "SDCCGRES", "EDUCDR04", "INCCGHH", "HWTCGBMI", 
                "PACCDPAI", "HCUC_1AA", "GENC_07", "SMKCDSTY", 
                "ALCCDTYP", "FVCCDTOT", "CCCC_071", "CCCC_91B",
                "CCCC_101", "GEOCGPRV", "WTSC_M")
cc21 <- c2[var.names2]
dim(cc21)
table(cc21$CCCC_051)
```

##### Cycle 3.1

```{r data3, warning=FALSE, message=FALSE, cache = TRUE}
var.names3 <- c("CCCE_121", "CCCE_051", "CCCE_05A", "DHHEGAGE", 
                "DHHE_SEX", "DHHEGMS", "SDCEGCGT", "SDCEFIMM", 
                "SDCEGRES", "EDUEDR04", "INCEGHH", "HWTEGBMI", 
                "PACEDPAI", "HCUE_1AA", "GENE_07", "SMKEDSTY", 
                "ALCEDTYP", "FVCEDTOT","CCCE_071", "CCCE_91F", 
                "CCCE_101", "GEOEGPRV", "WTSE_M")
cc31 <- c3[var.names3]
dim(cc31)
table(cc31$CCCE_051)
```

### Making variable names the same

We now create a new set of more readable and consistent variable names.

```{r data4, warning=FALSE, message=FALSE, cache = TRUE}
new.var.names <- c("CVD", "arthritis", "arthritis.kind", "age", 
                   "sex", "married", "race", "immigration", 
                   "recent.immigrant", "edu", "income", "bmi", 
                   "phyact", "doctor", "stress", "smoke",
                   "drink", "fruit", "bp", "copd", 
                   "diab",  "province", "weight")
cbind(new.var.names, var.names1, var.names2, var.names3)
names(cc11) <- names(cc21) <- names(cc31) <- new.var.names

table(cc11$arthritis)
table(cc21$arthritis)
table(cc31$arthritis)

cc11$cycle <- 11
cc21$cycle <- 21
cc31$cycle <- 31
```

### Appending

We now combine the data frames `cc11`, `cc21`, and `cc31` by stacking them on top of each other.

```{r append, warning=FALSE, message=FALSE, cache = TRUE}
cc123a <- rbind(cc11,cc21,cc31)
dim(cc123a)
names(cc123a)
cc123a$ID <- 1:nrow(cc123a)
```

### Variables

#### Sampling weight

We use the `summary` function to provide basic statistics (like mean, median, min, max, etc.) for the `weight` column in the data frame `cc123a`.

```{r summ, warning=FALSE, message=FALSE, cache = TRUE}
summary(cc123a$weight)
```

#### Exposure

This following chunk creates frequency tables for the 'arthritis' and 'arthritis.kind' columns.

```{r tab0, warning=FALSE, message=FALSE, cache = TRUE}
table(cc123a$arthritis)
table(cc123a$arthritis.kind)
sum(cc123a$arthritis=="NO")
sum(cc123a$arthritis.kind=="NOT APPLICABLE")
```

We create the exposure variable with exposure status vs controls.

```{r recodeOA, warning=FALSE, message=FALSE, cache = TRUE}
c123sub1 <- cc123a
c123sub1$arthritis.kind <- car::recode(c123sub1$arthritis.kind, 
                            "'OSTEOARTHRITIS'='OSTEOARTHRITIS';
                            'NOT APPLICABLE' = 'NOT APPLICABLE';
                            else = NA",  
                            as.factor = FALSE)
table(c123sub1$arthritis.kind, useNA = "always")
```

```{r recode1, warning=FALSE, message=FALSE, cache = TRUE}
# c123sub1 <- subset(cc123a, arthritis.kind == "OSTEOARTHRITIS" | 
#                      arthritis.kind == "NOT APPLICABLE" )
# dim(c123sub1)
table(c123sub1$arthritis.kind)
table(c123sub1$arthritis)
require(car)
c123sub1$arthritis.kind <- car::recode(c123sub1$arthritis.kind, 
                                  "'OSTEOARTHRITIS'='OA';
                         'NOT APPLICABLE'='Control';
                         else=NA", as.factor = FALSE)
table(c123sub1$arthritis.kind, useNA = "always")
c123sub1$OA <- c123sub1$arthritis.kind
c123sub1$arthritis.kind <- NULL
c123sub1$arthritis <- NULL
dim(c123sub1)
dim(c123sub1)[1]-dim(cc123a)[1]
```

#### Outcome

We create the outcome variable.

```{r recodecvd, warning=FALSE, message=FALSE, cache = TRUE}
c123sub2 <- c123sub1
table(c123sub2$CVD)
c123sub2$CVD <- car::recode(c123sub2$CVD, 
                            "'YES'='YES';
                            'NO' = 'NO';
                            else = NA",  
                            as.factor = FALSE)
```

```{r recodecvd3, warning=FALSE, message=FALSE, cache = TRUE}
# table(c123sub1$CVD)
# c123sub2 <- subset(c123sub1, CVD == "YES" | CVD == "NO")
# table(c123sub2$CVD)
dim(c123sub2)
c123sub2$CVD <- car::recode(c123sub2$CVD, 
                       "'YES'='event';
                       'NO'='no event';
                       else=NA",
                       as.factor = FALSE)
table(c123sub2$CVD, useNA = "always")
dim(c123sub2)
dim(c123sub2)[1]-dim(c123sub1)[1]
```

#### Covariates

##### Age

Recodes the 'age' column into broader age categories.

```{r recode4, warning=FALSE, message=FALSE, cache = TRUE}
c123sub2$age <- car::recode(c123sub2$age, 
                            "c('12 TO 14 YEARS','15 TO 19 YEARS', 
                            '15 TO 17 YEARS', '18 TO 19 YEARS')='teen';
                            c('20 TO 24 YEARS','25 TO 29 YEARS')=
                            '20-29 years'; 
                            c('30 TO 34 YEARS','35 TO 39 YEARS')=
                            '30-39 years';
                            c('40 TO 44 YEARS','45 TO 49 YEARS')=
                            '40-49 years';
                            c('50 TO 54 YEARS','55 TO 59 YEARS')=
                            '50-59 years';
                            c('60 TO 64 YEARS')='60-64 years';
                            else='65 years and over'",  
                            as.factor = FALSE)
table(c123sub2$age)
```

```{r recode5, warning=FALSE, message=FALSE, cache = TRUE}
c123sub3 <- c123sub2
# c123sub3$age[c123sub3$age == 'teen'] <- NA
# c123sub3$age[c123sub3$age == '65 years and over'] <- NA
```

```{r tabage, warning=FALSE, message=FALSE, cache = TRUE}
# dim(c123sub2)
# c123sub3 <- subset(c123sub2, age != 'teen' & age != '65 years and over')
table(c123sub3$age, useNA = "always")
dim(c123sub3)
```

##### Sex

Recodes 'sex' to 'Male' or 'Female'.

```{r recode6, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$sex)
c123sub3$sex <- car::recode(c123sub3$sex, 
                            "'MALE'='Male';
                            'FEMALE' = 'Female';
                            else = NA",  
                            as.factor = FALSE)
table(c123sub3$sex, useNA = "always")
```

##### Marital status

Recodes the 'married' column into two categories: 'not single' and 'single'.

```{r recode7, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$married)
c123sub3$married <- car::recode(c123sub3$married, 
                             "c('MARRIED', 'COMMON-LAW')='not single';
                             c('WIDOW/SEP/DIV', 'SINGLE', 
                             'SINGLE/NEVER MAR') = 'single';
                             else = NA",  
                             as.factor = FALSE)
table(c123sub3$married, useNA = "always")
```

##### Race/ethnicity

Recodes 'race' into 'White' and 'Non-white'.

```{r recode8, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$race)
c123sub3$race <- car::recode(c123sub3$race, 
                             "'WHITE'='White';
                             'VISIBLE MINORITY' = 'Non-white';
                             else = NA",  
                             as.factor = FALSE)
table(c123sub3$race, useNA = "always")
```

##### Immigration

Creates a new column for immigration status based on the 'recent.immigrant' column, then removes the original column.

```{r recode9, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$recent.immigrant)
c123sub3$immigrate <- car::recode(c123sub3$recent.immigrant,
                            "'0 TO 9 YEARS'='recent';
                            '10 YEARS OR MORE' = '> 10 years';
                            'NOT APPLICABLE' = 'not immigrant';
                            else = NA",
                            as.factor = FALSE)
table(c123sub3$immigrate, useNA = "always")
c123sub3$recent.immigrant <- NULL
c123sub3$immigration <- NULL
```

##### Education

Recode educational status into specified categories.

```{r recode10, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$edu)
c123sub3$edu <- car::recode(c123sub3$edu,
                            "'< THAN SECONDARY'='< 2ndary';
                            'SECONDARY GRAD.' = '2nd grad.';
                            'POST-SEC. GRAD.' = 'Post-2nd grad.';
                            'OTHER POST-SEC.' = 'Other 2nd grad.';
                            else = NA",
                            as.factor = FALSE)
table(c123sub3$edu, useNA = "always")
```

##### Income

Recodes income levels into broader categories.

```{r recode11, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$income)
# cycle 1.1 has: 'NO INCOME','LESS THAN 15,000'
# Other cycles have: 'NO OR <$15,000'
c123sub3$income <- car::recode(c123sub3$income, 
                               "c('NO OR <$15,000', 'NO INCOME',
                               'LESS THAN 15,000',
                               '$15,000-$29,999')='$29,999 or less';
                               '$30,000-$49,999' = '$30,000-$49,999';
                               '$50,000-$79,999' = '$50,000-$79,999';
                               '$80,000 OR MORE' = '$80,000 or more';
                               else = NA",  
                               as.factor = FALSE)
table(c123sub3$income, useNA = "always")
```

##### BMI

Converts 'bmi' column values to numerical data and also categorizes it based on the BMI value.

If you want to reuse the continuous variable later (usually a good idea in statistical sense), keep a second copy of the variable.

```{r recode12, warning=FALSE, message=FALSE, cache = TRUE}
# table(c123sub3$bmi)
sum(c123sub3$bmi=="NOT APPLICABLE")+
  sum(c123sub3$bmi=="REFUSAL")+
  sum(c123sub3$bmi=="NOT STATED")+
  sum(c123sub3$bmi=="DON'T KNOW")
c123sub3$bmi <- car::recode(c123sub3$bmi, 
                               'c("NOT APPLICABLE", "REFUSAL", 
                               "NOT STATED", "DON\'T KNOW")=NA',
                               as.factor = FALSE)
#table(c123sub3$bmi, useNA = "always")
c123sub3$bmi <- as.numeric(as.character(c123sub3$bmi))
summary(c123sub3$bmi)
c123sub3$bmi2 <- cut(c123sub3$bmi,
                       breaks = c(0,18.5,25,Inf),
                       right = TRUE,
                       labels = c("Underweight", 
                                  "healthy weight", 
                                  "Overweight"))
c123sub3$bmi <- c123sub3$bmi2
c123sub3$bmi2 <- NULL
table(c123sub3$bmi, useNA = "always")
```

##### Physical activity

Recodes physical activity levels.

```{r recode13, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$phyact)
c123sub3$phyact <- car::recode(c123sub3$phyact,
                               "'ACTIVE'='Active';
                               'MODERATE' = 'Moderate';
                               'INACTIVE' = 'Inactive';
                               else = NA",
                               as.factor = FALSE)
table(c123sub3$phyact, useNA = "always")
```

##### Doctor

Recodes whether someone has access to a doctor into 'Yes' or 'No'.

```{r recode14, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$doctor)
c123sub3$doctor <- car::recode(c123sub3$doctor, 
                            "'YES'='Yes';
                            'NO' = 'No';
                            else = NA",  
                            as.factor = FALSE)
table(c123sub3$doctor, useNA = "always")
```

##### Stress

Recodes stress levels into 'Not too stressed' or 'stressed'.

```{r recode16, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$stress)
c123sub3$stress <- car::recode(c123sub3$stress,
                              "c('NOT AT ALL','NOT VERY','A BIT')=
                              'Not too stressed';
                              c('QUITE A BIT','EXTREMELY') = 
                              'stressed';
                              else = NA",
                              as.factor = FALSE)
table(c123sub3$stress, useNA = "always")
```

##### Smoking

Recodes smoking status into 'Current smoker', 'Former smoker', or 'Never smoker'.

```{r recode17, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$smoke)
c123sub3$smoke <- car::recode(c123sub3$smoke,
                              "c('DAILY','OCCASIONAL',
                              'ALWAYS OCCASION.')='Current smoker';
                              c('FORMER DAILY','FORMER OCCASION.',
                              'ALWAYS OCCASION.') = 'Former smoker';
                              'NEVER SMOKED' = 'Never smoker';
                              else = NA",
                              as.factor = FALSE)
table(c123sub3$smoke, useNA = "always")
```

##### Alcohol

Recodes drinking habits into 'Current drinker', 'Former driker', or 'Never drank'.

```{r recode18, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$drink)
c123sub3$drink <- car::recode(c123sub3$drink,
                              "c('REGULAR DRINKER',
                              'OCC. DRINKER')='Current drinker';
                              c('FORMER DRINKER') = 'Former driker';
                              'NEVER DRANK' = 'Never drank';
                              else = NA",
                              as.factor = FALSE)
table(c123sub3$drink, useNA = "always")
```

##### Fruit and vegetable consumption

Converts fruit and vegetable consumption to numerical values and then categorizes it.

If you want to reuse the continuous variable later (usually a good idea in statistical sense), keep a second copy of the variable.

```{r recode19, warning=FALSE, message=FALSE, cache = TRUE}
str(c123sub3$fruit)
#c123sub3$fruit.cont <- c123sub3$fruit
c123sub3$fruit2 <- as.numeric(as.character(c123sub3$fruit))
# Note: do not use as.numeric(c123sub3$fruit)
summary(c123sub3$fruit2)
c123sub3$fruit2 <- cut(c123sub3$fruit2,
                       breaks = c(0,3,6,Inf),
                       right = TRUE,
                       labels = c("0-3 daily serving",
                                  "4-6 daily serving",
                                  "6+ daily serving"))
table(c123sub3$fruit2, useNA = "always")
c123sub3$fruit <- c123sub3$fruit2
c123sub3$fruit2 <- NULL
```

##### Hypertension

The following chunk of code is concerned with recoding the blood pressure column. The code also shows the distribution of the values before and after recoding.

```{r recode20, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$bp)
c123sub3$bp <- car::recode(c123sub3$bp,
                           "'YES'='Yes';
                           'NO' = 'No';
                           else = NA",
                           as.factor = FALSE)
table(c123sub3$bp, useNA = "always")
```

##### COPD

The following chunk of code is concerned with recoding the COPD column.

```{r recode21, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$copd)
c123sub3$copd <- car::recode(c123sub3$copd,
                             "'YES'='Yes';
                             'NO' = 'No';
                             else = NA",
                             as.factor = FALSE)
table(c123sub3$copd, useNA = "always")
```

##### Diabetes

The following chunk of code is concerned with recoding the diabetes column.

```{r recode22, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$diab)
c123sub3$diab <- car::recode(c123sub3$diab,
                             "'YES'='Yes';
                             'NO' = 'No';
                             else = NA",
                             as.factor = FALSE)
table(c123sub3$diab, useNA = "always")
```

##### North

This section is concerned with recoding the province column. It groups provinces into "North" and "South" based on their names.

-   Note that the category names might not exactly match with the data dictionary:
    -   Particularly note two versions of `QUEBEC`
    -   `PEI` was written in short
    -   None in
        -   DON'T KNOW
        -   REFUSAL
        -   NOT STATED
        -   NOT APPLICABLE

```{r recode23, warning=FALSE, message=FALSE, cache = TRUE}
c123sub3$province.check <- c123sub3$province
table(c123sub3$province)
c123sub3$province <- car::recode(c123sub3$province,
                              "c('YUKON/NWT/NUNAVT','YUKON/NWT/NUNA.')=
                              'North'; else = 'South'",
                              as.factor = FALSE)
table(c123sub3$province, useNA = "always")
```

##### Dimension testing

This chunk verifies the dimensions of the modified data frames and performs some other operations like setting factors.

```{r dims2, warning=FALSE, message=FALSE, cache = TRUE}
save.c123sub3 <- c123sub3
names(c123sub3)
# c123sub3$phyact <- NULL
# c123sub3$stress <- NULL
# c123sub3$fruit <- NULL
# c123sub3$copd <- NULL
# c123sub3$province.check <- NULL
dim(c123sub3)
dim(c123sub2)[1]-dim(c123sub3)[1]
analytic <- c123sub3
dim(analytic)
analytic$cycle <- as.factor(analytic$cycle)
names(analytic)
```

### Saving data

We save the data for future use:

```{r save, warning=FALSE, message=FALSE, cache = TRUE}
save(cc123a, analytic, file = "Data/surveydata/cchs123.RData")
```

### Video content (optional)

::: callout-tip
For those who prefer a video walkthrough, feel free to watch the video below, which offers a description of an earlier version of the above content.
:::

::: {style="position: relative; padding-bottom: 56.25%; height: 0; overflow: hidden;"}
<iframe src="https://www.youtube.com/embed/OWVV1O3iYmA" style="position: absolute; top: 0; left: 0; width: 100%; height: 100%; border:0;" allowfullscreen>

</iframe>
:::

## References
