## Creating dataset {.unnumbered}

```{r setup, warning=FALSE, message=FALSE, cache=TRUE}
# Load required packages
library(survey)
library(knitr)
library(car)
```

### Load data

```{r data, warning=FALSE, message=FALSE, cache = FALSE}
# CCHS 1.1
load("Data/surveydata/cchsc1.RData")

# CCHS 2.1
load("Data/surveydata/cchsc2.RData")

# CCHS 3.1
load("Data/surveydata/cchsc3.RData")

# objects loaded
ls()
```

```{r dim, warning=FALSE, message=FALSE, cache = TRUE}
# Dimensions
dim(c1)
dim(c2)
dim(c3)
```

### Subset the data (subset variables)

#### Understand variable coding

Ref:

-   page 60 in CCHS 3.1 guide [@statistics2005canadian]

-   link for variable list: [@varlist]

| **Variable Concept**                                              | **CCHS 1.1** | **CCHS 2.1** | **CCHS 3.1** |
|------------------------|----------------|----------------|----------------|
| **Has heart disease**                                             | CCCA_121     | CCCC_121     | CCCE_121     |
| **Has arthritis or rheumatism**                                   | CCCA_051     | CCCC_051     | CCCE_051     |
| **Kind of arthritis**                                             | CCCA_05A     | CCCC_05A     | CCCE_05A     |
| **Age**                                                           | DHHAGAGE     | DHHCGAGE     | DHHEGAGE     |
| **Sex**                                                           | DHHA_SEX     | DHHC_SEX     | DHHE_SEX     |
| **Marital Status**                                                | DHHAGMS      | DHHCGMS      | DHHEGMS      |
| **Cultural / racial origin**                                      | SDCAGRAC     | SDCCGRAC     | SDCEGCGT     |
| **Immigrant status**                                              | SDCAFIMM     | SDCCFIMM     | SDCEFIMM     |
| **Length of time in Canada since immigration**                    | SDCAGRES     | SDCCGRES     | SDCEGRES     |
| **Highest level of education - respondent**                       | EDUADR04     | EDUCDR04     | EDUEDR04     |
| **Total household income from all sources**                       | INCAGHH      | INCCGHH      | INCEGHH      |
| **Body mass index**                                               | HWTAGBMI     | HWTCGBMI     | HWTEGBMI     |
| **Physical activity index**                                       | PACADPAI     | PACCDPAI     | PACEDPAI     |
| **Has a regular medical doctor**                                  | TWDA_5       | HCUC_1AA     | HCUE_1AA     |
| **Self-perceived stress**                                         | GENA_07      | GENC_07      | GENE_07      |
| **Type of smoker**                                                | SMKADSTY     | SMKCDSTY     | SMKEDSTY     |
| **Type of drinker**                                               | ALCADTYP     | ALCCDTYP     | ALCEDTYP     |
| **Daily consumption - total fruits and vegetables**               | FVCADTOT     | FVCCDTOT     | FVCEDTOT     |
| **Has high blood pressure**                                       | CCCA_071     | CCCC_071     | CCCE_071     |
| **Has emphysema or chronic obstructive pulmonary disease (COPD)** | CCCA_91B     | CCCC_91B     | CCCE_91F     |
| **Has diabetes**                                                  | CCCA_101     | CCCC_101     | CCCE_101     |
| **Province**                                                      | GEOAGPRV     | GEOCGPRV     | GEOEGPRV     |
| **Sampling weight - master weight**                               | WTSAM        | WTSC_M       | WTSE_M       |

We only focus on CCHS 3.1 content here, but need to check all other cycles as well.

Most variables have: Universe \`All respondents'.

Below are some exceptions:

-   CCCE_05A universe: (Kind of arthritis / rheumatism)

-   Respondents who answered CCCE_051 = (1, 7 or 8) or CCCE_011 = 8

-   CCCE_051: All respondents

-   CCCE_011: All respondents

-   SDCEGRES universe: (Length of time in Canada since immigration)

    -   Respondents who answered SDCE_2 = (2, 7 or 8) or SDCE_1 = (97 or 98)
    -   SDCE_2 doesn't exist!
    -   (master file variable; not available in PUMF)
    -   Public Use Microdata File (PUMF)
    -   SDCE_1 doesn't exist!
    -   (master file variable; not available in PUMF)

-   HWTEGBMI universe: (Body Mass Index (BMI) / self-report)

    -   All respondents excluding pregnant women (MAME_037 = 1)
        -   MAME_037 doesn't exist!
        -   (master file variable; not available in PUMF)

-   GENE_07 universe: (Self-perceived stress)

    -   Respondents aged 15 and over

-   CCCE_91F universe: (Has chronic obstructive pulmonary disease)

    -   Respondents aged 30 and over

-   FVCEDTOT universe: (Daily consumption - total fruits and vegetables)

    -   Respondents with FVCEFOPT = 1
        -   FVCEFOPT: Optional module: Fruit and vegetable consumption - (F)

Ref:

-   page 66 in CCHS 3.1 guide [@statistics2005canadian]

-   Potential problematic variables:

    -   Self-perceived stress
    -   Has chronic obstructive pulmonary disease / copd
    -   Daily consumption - total fruits and vegetables

We will make decisions about these variables later: for now, let's keep them.

#### Restrict the dataset with variables of interest only

##### Cycle 1.1

```{r data1, warning=FALSE, message=FALSE, cache = TRUE}
var.names1 <- c("CCCA_121", 
                "CCCA_051", 
                "CCCA_05A", 
                "DHHAGAGE", 
                "DHHA_SEX", 
                "DHHAGMS", 
                "SDCAGRAC", 
                "SDCAFIMM", 
                "SDCAGRES", 
                "EDUADR04", 
                "INCAGHH", 
                "HWTAGBMI", 
                "PACADPAI", 
                "TWDA_5", 
                "GENA_07", 
                "SMKADSTY", 
                "ALCADTYP", 
                "FVCADTOT", 
                "CCCA_071",
                "CCCA_91B",
                "CCCA_101",
                "GEOAGPRV",
                "WTSAM")
cc11 <- c1[var.names1]
dim(cc11)
table(cc11$CCCA_051)
```

##### Cycle 2.1

```{r data2, warning=FALSE, message=FALSE, cache = TRUE}
var.names2 <- c("CCCC_121", 
                "CCCC_051", 
                "CCCC_05A", 
                "DHHCGAGE", 
                "DHHC_SEX", 
                "DHHCGMS", 
                "SDCCGRAC", 
                "SDCCFIMM", 
                "SDCCGRES", 
                "EDUCDR04", 
                "INCCGHH", 
                "HWTCGBMI", 
                "PACCDPAI", 
                "HCUC_1AA", 
                "GENC_07", 
                "SMKCDSTY", 
                "ALCCDTYP", 
                "FVCCDTOT", 
                "CCCC_071",
                "CCCC_91B",
                "CCCC_101",
                "GEOCGPRV",
                "WTSC_M")
cc21 <- c2[var.names2]
dim(cc21)
table(cc21$CCCC_051)
```

##### Cycle 3.1

```{r data3, warning=FALSE, message=FALSE, cache = TRUE}
var.names3 <- c("CCCE_121", 
                "CCCE_051", 
                "CCCE_05A", 
                "DHHEGAGE", 
                "DHHE_SEX", 
                "DHHEGMS", 
                "SDCEGCGT", 
                "SDCEFIMM", 
                "SDCEGRES", 
                "EDUEDR04", 
                "INCEGHH", 
                "HWTEGBMI", 
                "PACEDPAI", 
                "HCUE_1AA", 
                "GENE_07", 
                "SMKEDSTY", 
                "ALCEDTYP", 
                "FVCEDTOT", 
                "CCCE_071", 
                "CCCE_91F", 
                "CCCE_101", 
                "GEOEGPRV", 
                "WTSE_M")
cc31 <- c3[var.names3]
dim(cc31)
table(cc31$CCCE_051)
```

### Making variable names the same

```{r data4, warning=FALSE, message=FALSE, cache = TRUE}
new.var.names <- c("CVD", 
                   "arthritis", 
                   "arthritis.kind", 
                   "age", 
                   "sex", 
                   "married", 
                   "race", 
                   "immigration", 
                   "recent.immigrant", 
                   "edu", 
                   "income", 
                   "bmi", 
                   "phyact", 
                   "doctor", 
                   "stress", 
                   "smoke", 
                   "drink", 
                   "fruit", 
                   "bp", 
                   "copd", 
                   "diab",  
                   "province", 
                   "weight")
cbind(new.var.names, var.names1, var.names2, var.names3)
names(cc11) <- names(cc21) <- names(cc31) <- new.var.names

table(cc11$arthritis)
table(cc21$arthritis)
table(cc31$arthritis)

cc11$cycle <- 11
cc21$cycle <- 21
cc31$cycle <- 31
```

### Appending

```{r append, warning=FALSE, message=FALSE, cache = TRUE}
cc123a <- rbind(cc11,cc21,cc31)
dim(cc123a)
names(cc123a)
cc123a$ID <- 1:nrow(cc123a)
```

### Variables

#### Sampling weight

```{r summ, warning=FALSE, message=FALSE, cache = TRUE}
summary(cc123a$weight)
```

#### Exposure

```{r tab0, warning=FALSE, message=FALSE, cache = TRUE}
table(cc123a$arthritis)
table(cc123a$arthritis.kind)
sum(cc123a$arthritis=="NO")
sum(cc123a$arthritis.kind=="NOT APPLICABLE")
```

```{r recodeOA, warning=FALSE, message=FALSE, cache = TRUE}
c123sub1 <- cc123a
c123sub1$arthritis.kind <- car::recode(c123sub1$arthritis.kind, 
                            "'OSTEOARTHRITIS'='OSTEOARTHRITIS';
                            'NOT APPLICABLE' = 'NOT APPLICABLE';
                            else = NA",  
                            as.factor = FALSE)
table(c123sub1$arthritis.kind, useNA = "always")
```

```{r recode1, warning=FALSE, message=FALSE, cache = TRUE}
# c123sub1 <- subset(cc123a, arthritis.kind == "OSTEOARTHRITIS" | 
#                      arthritis.kind == "NOT APPLICABLE" )
# dim(c123sub1)
table(c123sub1$arthritis.kind)
table(c123sub1$arthritis)
require(car)
c123sub1$arthritis.kind <- car::recode(c123sub1$arthritis.kind, 
                                  "'OSTEOARTHRITIS'='OA';
                         'NOT APPLICABLE'='Control';
                         else=NA", as.factor = FALSE)
table(c123sub1$arthritis.kind, useNA = "always")
c123sub1$OA <- c123sub1$arthritis.kind
c123sub1$arthritis.kind <- NULL
c123sub1$arthritis <- NULL
dim(c123sub1)
dim(c123sub1)[1]-dim(cc123a)[1]
```

#### Outcome

```{r recodecvd, warning=FALSE, message=FALSE, cache = TRUE}
c123sub2 <- c123sub1
table(c123sub2$CVD)
c123sub2$CVD <- car::recode(c123sub2$CVD, 
                            "'YES'='YES';
                            'NO' = 'NO';
                            else = NA",  
                            as.factor = FALSE)
```

```{r recodecvd3, warning=FALSE, message=FALSE, cache = TRUE}
# table(c123sub1$CVD)
# c123sub2 <- subset(c123sub1, CVD == "YES" | CVD == "NO")
# table(c123sub2$CVD)
dim(c123sub2)
c123sub2$CVD <- car::recode(c123sub2$CVD, 
                       "'YES'='event';
                       'NO'='no event';
                       else=NA",
                       as.factor = FALSE)
table(c123sub2$CVD, useNA = "always")
dim(c123sub2)
dim(c123sub2)[1]-dim(c123sub1)[1]
```

#### Covariates

##### age

```{r recode4, warning=FALSE, message=FALSE, cache = TRUE}
c123sub2$age <- car::recode(c123sub2$age, 
                       "c('12 TO 14 YEARS','15 TO 19 YEARS',
                       '15 TO 17 YEARS', '18 TO 19 YEARS')='teen';
                       c('20 TO 24 YEARS','25 TO 29 YEARS')='20-29 years';
                       c('30 TO 34 YEARS','35 TO 39 YEARS')='30-39 years';
                       c('40 TO 44 YEARS','45 TO 49 YEARS')='40-49 years';
                       c('50 TO 54 YEARS','55 TO 59 YEARS')='50-59 years';
                       c('60 TO 64 YEARS')='60-64 years';   
                       else='65 years and over'",
                       as.factor = FALSE)
table(c123sub2$age)
```

```{r recode5, warning=FALSE, message=FALSE, cache = TRUE}
c123sub3 <- c123sub2
# c123sub3$age[c123sub3$age == 'teen'] <- NA
# c123sub3$age[c123sub3$age == '65 years and over'] <- NA
```

```{r tabage, warning=FALSE, message=FALSE, cache = TRUE}
# dim(c123sub2)
# c123sub3 <- subset(c123sub2, age != 'teen' & age != '65 years and over')
table(c123sub3$age, useNA = "always")
dim(c123sub3)
```

##### sex

```{r recode6, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$sex)
c123sub3$sex <- car::recode(c123sub3$sex, 
                            "'MALE'='Male';
                            'FEMALE' = 'Female';
                            else = NA",  
                            as.factor = FALSE)
table(c123sub3$sex, useNA = "always")
```

##### marital status

```{r recode7, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$married)
c123sub3$married <- car::recode(c123sub3$married, 
                             "c('MARRIED', 'COMMON-LAW')='not single';
                             c('WIDOW/SEP/DIV', 'SINGLE', 'SINGLE/NEVER MAR') = 'single';
                             else = NA",  
                             as.factor = FALSE)
table(c123sub3$married, useNA = "always")
```

##### race

```{r recode8, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$race)
c123sub3$race <- car::recode(c123sub3$race, 
                             "'WHITE'='White';
                             'VISIBLE MINORITY' = 'Non-white';
                             else = NA",  
                             as.factor = FALSE)
table(c123sub3$race, useNA = "always")
```

##### immigration

```{r recode9, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$recent.immigrant)
c123sub3$immigrate <- car::recode(c123sub3$recent.immigrant,
                            "'0 TO 9 YEARS'='recent';
                            '10 YEARS OR MORE' = '> 10 years';
                            'NOT APPLICABLE' = 'not immigrant';
                            else = NA",
                            as.factor = FALSE)
table(c123sub3$immigrate, useNA = "always")
c123sub3$recent.immigrant <- NULL
c123sub3$immigration <- NULL
```

##### education

```{r recode10, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$edu)
c123sub3$edu <- car::recode(c123sub3$edu,
                            "'< THAN SECONDARY'='< 2ndary';
                            'SECONDARY GRAD.' = '2nd grad.';
                            'POST-SEC. GRAD.' = 'Post-2nd grad.';
                            'OTHER POST-SEC.' = 'Other 2nd grad.';
                            else = NA",
                            as.factor = FALSE)
table(c123sub3$edu, useNA = "always")
```

##### income

```{r recode11, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$income)
# cycle 1.1 has: 'NO INCOME','LESS THAN 15,000'
# Other cycles have: 'NO OR <$15,000'
c123sub3$income <- car::recode(c123sub3$income, 
                               "c('NO OR <$15,000', 'NO INCOME','LESS THAN 15,000',
                               '$15,000-$29,999')='$29,999 or less';
                               '$30,000-$49,999' = '$30,000-$49,999';
                               '$50,000-$79,999' = '$50,000-$79,999';
                               '$80,000 OR MORE' = '$80,000 or more';
                               else = NA",  
                               as.factor = FALSE)
table(c123sub3$income, useNA = "always")
```

##### BMI

If you want to reuse the continuous variable later (usually a good idea in statistical sense), keep a second copy of the variable.

```{r recode12, warning=FALSE, message=FALSE, cache = TRUE}
# table(c123sub3$bmi)
sum(c123sub3$bmi=="NOT APPLICABLE")+
  sum(c123sub3$bmi=="REFUSAL")+
  sum(c123sub3$bmi=="NOT STATED")+
  sum(c123sub3$bmi=="DON'T KNOW")
c123sub3$bmi <- car::recode(c123sub3$bmi, 
                               'c("NOT APPLICABLE", "REFUSAL", "NOT STATED", "DON\'T KNOW")=NA',
                               as.factor = FALSE)
#table(c123sub3$bmi, useNA = "always")
c123sub3$bmi <- as.numeric(as.character(c123sub3$bmi))
summary(c123sub3$bmi)
c123sub3$bmi2 <- cut(c123sub3$bmi,
                       breaks = c(0,18.5,25,Inf),
                       right = TRUE,
                       labels = c("Underweight", 
                                  "healthy weight", 
                                  "Overweight"))
c123sub3$bmi <- c123sub3$bmi2
c123sub3$bmi2 <- NULL
table(c123sub3$bmi, useNA = "always")
```

##### physical activity

```{r recode13, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$phyact)
c123sub3$phyact <- car::recode(c123sub3$phyact,
                               "'ACTIVE'='Active';
                               'MODERATE' = 'Moderate';
                               'INACTIVE' = 'Inactive';
                               else = NA",
                               as.factor = FALSE)
table(c123sub3$phyact, useNA = "always")
```

##### doctor

```{r recode14, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$doctor)
c123sub3$doctor <- car::recode(c123sub3$doctor, 
                            "'YES'='Yes';
                            'NO' = 'No';
                            else = NA",  
                            as.factor = FALSE)
table(c123sub3$doctor, useNA = "always")
```

##### stress

```{r recode16, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$stress)
c123sub3$stress <- car::recode(c123sub3$stress,
                              "c('NOT AT ALL','NOT VERY','A BIT')='Not too stressed';
                              c('QUITE A BIT','EXTREMELY') = 'stressed';
                              else = NA",
                              as.factor = FALSE)
table(c123sub3$stress, useNA = "always")
```

##### smoking

```{r recode17, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$smoke)
c123sub3$smoke <- car::recode(c123sub3$smoke,
                              "c('DAILY','OCCASIONAL',
                              'ALWAYS OCCASION.')='Current smoker';
                              c('FORMER DAILY','FORMER OCCASION.',
                              'ALWAYS OCCASION.') = 'Former smoker';
                              'NEVER SMOKED' = 'Never smoker';
                              else = NA",
                              as.factor = FALSE)
table(c123sub3$smoke, useNA = "always")
```

##### alcohol

```{r recode18, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$drink)
c123sub3$drink <- car::recode(c123sub3$drink,
                              "c('REGULAR DRINKER',
                              'OCC. DRINKER')='Current drinker';
                              c('FORMER DRINKER') = 'Former driker';
                              'NEVER DRANK' = 'Never drank';
                              else = NA",
                              as.factor = FALSE)
table(c123sub3$drink, useNA = "always")
```

##### fruit and vegetable consumption

If you want to reuse the continuous variable later (usually a good idea in statistical sense), keep a second copy of the variable.

```{r recode19, warning=FALSE, message=FALSE, cache = TRUE}
str(c123sub3$fruit)
#c123sub3$fruit.cont <- c123sub3$fruit
c123sub3$fruit2 <- as.numeric(as.character(c123sub3$fruit))
# Note: do not use as.numeric(c123sub3$fruit)
summary(c123sub3$fruit2)
c123sub3$fruit2 <- cut(c123sub3$fruit2,
                       breaks = c(0,3,6,Inf),
                       right = TRUE,
                       labels = c("0-3 daily serving",
                                  "4-6 daily serving",
                                  "6+ daily serving"))
table(c123sub3$fruit2, useNA = "always")
c123sub3$fruit <- c123sub3$fruit2
c123sub3$fruit2 <- NULL
```

##### hypertension

```{r recode20, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$bp)
c123sub3$bp <- car::recode(c123sub3$bp,
                           "'YES'='Yes';
                           'NO' = 'No';
                           else = NA",
                           as.factor = FALSE)
table(c123sub3$bp, useNA = "always")
```

##### COPD

```{r recode21, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$copd)
c123sub3$copd <- car::recode(c123sub3$copd,
                             "'YES'='Yes';
                             'NO' = 'No';
                             else = NA",
                             as.factor = FALSE)
table(c123sub3$copd, useNA = "always")
```

##### Diabetes

```{r recode22, warning=FALSE, message=FALSE, cache = TRUE}
table(c123sub3$diab)
c123sub3$diab <- car::recode(c123sub3$diab,
                             "'YES'='Yes';
                             'NO' = 'No';
                             else = NA",
                             as.factor = FALSE)
table(c123sub3$diab, useNA = "always")
```

##### North

-   Note that the category names might not exactly match with the data dictionary:
    -   Particularly note two versions of `QUEBEC`
    -   `PEI` was written in short
    -   None in
        -   DON'T KNOW
        -   REFUSAL
        -   NOT STATED
        -   NOT APPLICABLE

```{r recode23, warning=FALSE, message=FALSE, cache = TRUE}
c123sub3$province.check <- c123sub3$province
table(c123sub3$province)
# c123sub3$province <- car::recode(c123sub3$province,
#                               "c('YUKON/NWT/NUNAVT','YUKON/NWT/NUNA.')='North';
#                               c('NEWFOUNDLAND', 'PEI', 'NOVA SCOTIA', 'NEW BRUNSWICK', 
# 'QU?BEC', 'ONTARIO', 'MANITOBA', 'SASKATCHEWAN', 'ALBERTA', 
# 'BRITISH COLUMBIA', 'QUEBEC', 'NFLD & LAB.') = 'South';
#                               else = NA",
#                               as.factor = FALSE)
c123sub3$province <- car::recode(c123sub3$province,
                              "c('YUKON/NWT/NUNAVT','YUKON/NWT/NUNA.')='North';
                              else = 'South'",
                              as.factor = FALSE)
table(c123sub3$province, useNA = "always")
```

##### Dimension testing

```{r dims2, warning=FALSE, message=FALSE, cache = TRUE}
save.c123sub3 <- c123sub3
names(c123sub3)
# c123sub3$phyact <- NULL
# c123sub3$stress <- NULL
# c123sub3$fruit <- NULL
# c123sub3$copd <- NULL
# c123sub3$province.check <- NULL
dim(c123sub3)
dim(c123sub2)[1]-dim(c123sub3)[1]
analytic <- c123sub3
dim(analytic)
analytic$cycle <- as.factor(analytic$cycle)
names(analytic)
```

### Saving data

```{r save, warning=FALSE, message=FALSE, cache = TRUE}
save(cc123a, analytic, file = "Data/surveydata/cchs123.RData")
```
